# Syllabus

```{r  include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gt)
```

This is the _draft_ syllabus for the Data Visualization (STAT 2430) course in Winter 2021 at Dalhousie University.

## Course Description

Data visualization is the art and science of turning data into readable graphics. We will explore
how to design and create data visualizations.
This exploration will include both the principles and
techniques of data visualization. Students will learn the value of visualization, specific techniques and understand
how to select the best visualization method.

## Course Prerequisites

At least one MATH or STAT course at or above the 1000 level. No experience with R or computer programming is required, but a desire to learn about both is essential.  

## Course Objectives and Learning Outcomes

By the end of the course you will be familiar with several aspects of data visualzation.

*Visual impact and aesthetic aspects of graphics*

- Understand the relationship between the structure of your data and the perceptual features of your graphics.
- Describe aesthetic aspects of good plots. 
- Use length, shape, size, colour, and other features to effectively display data and enable comparitive visual interpretation.

*Visualization as communication*

- Visualization is a visual language for communication, which should be accompanied by written interpretations
- Communication effectiveness should be evaluated by peer-review and critical, constructive feedback
- Visualizations should be developed through an iterative process akin to editing text, including the process of refining plots to highlight key features of the data, labeling items of interest, annotating plots, and changing overall appearance
- Visualizations can be presented in different formats for different of audiences and different communications goals

*Computing skills*

- Learn the basics of using R, Rstudio, several R add-on packages, and git
- Read data in several different formats into R
- Create graphs with ggplot2 for continuous and categorical variables with informative legends.
- Add error bars, linear models (OLS, robust regression, polynomials, quantile regression), smooths (LOESS, GAM), labels, and other annotations to a graph
- Create small-multiple (facetted) plots.
- Use the principles of [tidy data](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html) tables to facilitate transformation and analysis of data. Reshape data to make it tidy.
- Summarize and transform data using dplyr
- Reshape data using pivots
- Create maps and some alternatives for presenting spatial data.
- Write reproducible documents with R markdown to document your analysis process and present your results
- Distribute data, code, and results using git and github
- Access and interpret help resources (built-in help, vignettes, web pages, online discussion forums, blogs).
- Develop skills for independently learning new data visualization methods and software

*Statistical models*

- Use a variety of modeling techniques such as LOESS, OLS, robust regression, polynomial regression, and quantile regression
- Learn how to extract model information to compare different statistical models
- Use principal component analysis (PCA) to reduce the dimensionality of complex datasets, increasing interpretability while minimizing information loss
- Use multidimensional scaling (MDS) to visualize and compare similarities and dissimilarities between variables
- Divide observations into homogeneous and distinct groups using K-means

## Course materials
 
Course notes:

- Andrew J Irwin (2021) [Data visualization](https://andrewirwin.github.io/data-visualization/) _Will be revised throughout the course._

Textbook: 

- Kieran Healy (2018) [Data Visualization: A Practical Introduction](https://socviz.co). Princeton University Press. 

Supplemental textbooks: 

- Garret Grolemund & Hadley Wickham (2017) [R for data science](https://r4ds.had.co.nz/). O'Reilly. 
- Claus O. Wilke (2019) [Fundamentals of data visualization](https://clauswilke.com/dataviz/). O'Reilly.

## Course structure

Each week will be structured around the following components:

- Recorded mini-lectures for each lesson (linked to course notes and slides)
- Reading course notes and excerpts from the textbooks linked to each lesson
- Interactive discussion of the week's lessons (Tuesday, 8:30 am Atlantic)
- Live coding tutorial / lab (linked to materials created before and during the meeting, Thursday 8:30 am Atlantic)
- Opportunities to develop and demonstrate your knowledge (tasks, assignments, and a term project)

If you have questions about the course material, please submit them on the following [discussion forum](https://piazza.com/dal.ca/winter2021/stat2430/home). If you have questions specific to your situation that should not be shared with the class, please send me an [email](a.irwin@dal.ca). Please only use email for communication that should be private, as general questions are best asked in public on the discussion forum where everyone in the class can benefit from the exchange. Please feel free to use the discussion forum to talk with other classmates and help each other.

Our first week of class does not meet on Tuesday but does meet on Thursday. This first meeting will be an introduction to the course and interactive discussion.

We will use Collaborate Ultra accessed through [Brightspace](https://dal.brightspace.com/d2l/home/159870) for the interactive sessions. Generally these will be recorded so that you can watch them later if you are unable to participate. You are strongly encouraged to come to these meetings. The interactive sessions are the primary forum for asking questions and discussing course material. These regular meetings will help give your work on the course structure. I strongly encourage you to "raise your hand" during the discussion and ask questions using your voice. Some students prefer to use the chat to ask questions, and that is fine with me, but from my perspective it leads to a more disjointed and less useful interaction.

## Evaluation

Tasks are opportunities to test your understanding of the key concepts from each lesson and demonstrate you have developed basic profiency with some computing skills. Most lessons will have a task for you to complete.

Assignments combine skills from multiple lessons into a meaningful data visualization activity. These will require using skills from multiple tasks and applying the content of lessons in thoughtful and sometimes creative ways.

The final project is an opportunity to combine many of the skills learned in the course. You will explore, analyze, and present to a reader an analysis of a dataset of your choice. The proposal will select a dataset and describe some of your planned analyses. The report will be a polished document showing off many of the ideas of the course using this dataset.

* Tasks for most lessons (30%, roughly 2-3 per week, equally weighted)
* Assignments (40%, roughly 1 every two weeks, equally weighted)
* Term project, divided into three components
  * Proposal, due Friday 5 March, 5%
  * Oral presentation, due Thursday 1 April, 10%
  * Report, due Friday 9 April, 15%

Tasks are due on Friday of the week they are assigned, unless otherwise noted. Late work will be accepted without penalty until the work is graded and solutions are posted. After that, no later work will be accepted, but if an accommodation is requested and granted, the value of that work will be redistributed to other tasks or assignments. *If you anticipate not being able to submit the proposal or report on time, please contact me by email.*

Grades will be reported on [Brightspace](https://dal.brightspace.com/d2l/home/159870)

## Software

The statistical software R/RStudio and version control software git will be used in this course. No prior experience with
R/RStudio is assumed, and we'll take class time to learn the software.

* To download and install R got to [r-project.org](https://www.r-project.org/) and click on the link to [Download R](http://cran.r-project.org/mirrors.html)
* To download and install Rstudio, go to [Rstudio.com](https://www.rstudio.com/) and click on the link do [Download Rstudio](https://rstudio.com/products/rstudio/download/)
* To download and install git 
  * on Windows, go to [git-scm.org](https://git-scm.org) and click on the link to [download a version for Windows](https://git-scm.com/download/win)
  * on Macintosh, use the Terminal app and type 'xcode-select --install' to download and install git.

These tools can be installed on Linux computers as well. Contact me if you have trouble. If you have a Chromebook you can use all these tools through the cloud service [rstudio.cloud](https://rstudio.cloud/) Everyone should learn to use the cloud service as a backup in case of problems with setting up the software on their own computer.

R and Rstudio are available on Dalhousie computer labs, but the git version control software must be installed following the instructions for Windows computers above. Since all your user files are erased from lab computers when you log out, this process must be repeated on each login.

[Videos](https://dal.brightspace.com/d2l/le/content/159870/Home?itemIdentifier=D2L.LE.Content.ContentObject.ModuleCO-2083726) demonstrating how to install this software are on Brightspace.

## Letter grades

Your numerical grade will be converted to a 
[letter grade for the course](https://www.dal.ca/campus_life/academic-support/grades-and-student-records/grade-scale-and-definitions.html) using the Dalhousie Common Grade Scale.

First your numerical grade will be rounded up to the nearest integer, then it will be converted to a letter using this table.

```{r echo=FALSE}
grade_table <- tibble(
        letter = c("A+", "A", "A-", "B+", "B", "B-", "C+", "C", "C-", "D", "F"),
        min_grade = c(90, 85, 80, 77, 73, 70, 65, 60, 55, 50, 0),
        max_grade = c(100, min_grade[1:10]-1)
)
grade_table %>% unite(`Grade range`, min_grade, max_grade, sep="-") %>%
        rename(`Letter grade`=letter) %>% gt() %>% cols_align("center")
```

## Course Policies

Credit will not be given for assignments submitted after grading is complete or the solutions are posted. If you miss an assignment due to illness or other event outside your control, contact me for an accommodation. If an accomodation is granted, assignment weight will be shifted to the rest of the assignments. 

Students need to finish all assignments independently; collaboration or copying solutions from any other sources are not allowed and will be considered an academic offense for this course. Tasks are primarily designed to help you learn and you are encouraged to seek assistance from classmates, but work you submit must be your own. The final project and its components may be done in groups, but the work on the project should only be the work of members of the group.

## University Policies and Statements

This course is governed by the academic rules and regulations set forth in the University Calendar and by Senate

*Missed or Late Academic Requirements due to Student Absence*
As per Senate decision instructors may not require medical notes of students who must miss an academic requirement, including the final exam, for courses offered during fall or winter 2020-21 (until April 30, 2021). Information on regular policy, including the use of the Student Declaration of Absence can be found here: \url{https://www.dal.ca/dept/university_secretariat/policies/academic/missed-or-late-academic-requirements-due-to-student-absence.html. }


*Academic Integrity*
At Dalhousie University, we are guided in all of our work by the values of academic integrity: honesty, trust, fairness, responsibility and respect (The Center for Academic Integrity, Duke University, 1999). As a student, you are required to demonstrate these values in all of the work you do. The University provides policies and procedures that every member of the university community is required to follow to ensure academic integrity. [More information](https://www.dal.ca/dept/university_secretariat/academic-integrity.html)

*Accessibility*
The Advising and Access Services Centre is Dalhousie's centre of expertise for student accessibility and accommodation. The advising team works with students who request accommodation as a result of a disability, religious obligation, or any barrier related to any other characteristic protected under Human Rights legislation (Canada and Nova Scotia). [More information](https://www.dal.ca/campus_life/academic-support/accessibility.html)

*Student Code of Conduct*
Everyone at Dalhousie is expected to treat others with dignity and respect. The Code of Student Conduct allows Dalhousie to take disciplinary action if students don’t follow this community expectation. When appropriate, violations of the code can be resolved in a reasonable and informal manner—perhaps through a restorative justice process. If an informal resolution can’t be reached, or would be inappropriate, procedures exist for formal dispute resolution. [Code](https://www.dal.ca/dept/university_secretariat/policies/student-life/code-of-student-conduct.html)

*Diversity and Inclusion - Culture of Respect*
Every person at Dalhousie has a right to be respected and safe. We believe inclusiveness is fundamental to education. We stand for equality. Dalhousie is strengthened in our diversity. We are a respectful and inclusive community. We are committed to being a place where everyone feels welcome and supported, which is why our Strategic Direction prioritizes fostering a culture of diversity and inclusiveness. [Statement](http://www.dal.ca/cultureofrespect.html)

*Recognition of Mi’kmaq Territory*
Dalhousie University would like to acknowledge that the University is on Traditional Mi’kmaq Territory. The Elders in Residence program provides students with access to First Nations elders for guidance, counsel and support. Visit or e-mail the Indigenous Student Centre (1321 Edward St) (elders@dal.ca). [Information](https://www.dal.ca/campus_life/communities/indigenous.html)

* [Important Dates in the Academic Year (including add/drop dates)](https://www.dal.ca/academics/important_dates.html)
* [University Grading Practices](https://www.dal.ca/dept/university_secretariat/policies/academic/grading-practices-policy.html)

## Student Resources and Support

*Advising*: 
[General Advising](https://www.dal.ca/campus_life/academic-support/advising.html),
[Science Program Advisors](https://www.dal.ca/faculty/science/current-students/academic-advising.html),
[Indigenous Student Centre](https://www.dal.ca/campus_life/communities/indigenous.html),
[Black Advising Centre](https://www.dal.ca/campus_life/communities/black-student-advising.html),
[International Centre](https://www.dal.ca/campus_life/international-centre/current-students.html)

*Academic supports*: 
[Library](https://libraries.dal.ca/), 
[Writing Centre](https://www.dal.ca/campus_life/academic-support/writing-and-study-skills.html),
[Studying for Success](https://www.dal.ca/campus_life/academic-support/study-skills-and-tutoring.html),
[Copyright Office](https://libraries.dal.ca/services/copyright-office.html),
[Fair Dealing Guidelines](https://libraries.dal.ca/services/copyright-office/fair-dealing.html)

*Other supports and services*:
[Student Health & Wellness Centre](https://www.dal.ca/campus_life/health-and-wellness/services-support/student-health-and-wellness.html),
[Student Advocacy](https://dsu.ca/dsas),
[Ombudsperson](https://www.dal.ca/campus_life/safety-respect/student-rights-and-responsibilities/where-to-get-help/ombudsperson.html)

*Safety*:
[Biosafety](https://www.dal.ca/dept/safety/programs-services/biosafety.html),
[Chemical Safety](https://www.dal.ca/dept/safety/programs-services/chemical-safety.html),
[Radiation Safety](https://www.dal.ca/dept/safety/programs-services/radiation-safety.html),
[Scent‐Free Program](https://www.dal.ca/dept/safety/programs-services/occupational-safety/scent-free.html)
