# Detailed outline {#outline .unnumbered}

```{r include=FALSE}
library(tidyverse)  # fussy, but might be useful later on
library(knitr)
library(kableExtra)
library(readxl)
library(lubridate)
cal <- read_excel("schedule.xlsx")
```

Synchronous sessions will be held on Thursday starting at 8:35 and ending at 9:55 or earlier. You can find the link to these sessions on [Brightspace](https://dal.brightspace.com/d2l/le/content/159870/Home?itemIdentifier=D2L.LE.Content.ContentObject.ModuleCO-2083791). There is an [agenda](#sync) for each session.  Most sessions will be recorded and posted on Brightspace. The other synchronous time in the timetable (Tuesday starting at 8:35) will be used for drop-in office hours. 

Each lesson has written notes, a recorded video, slides, and links to textbook readings in the notes. Some students will prefer to begin with the video, others will prefer to begin with the written notes and slides. I suggest you learn which you prefer by experimentation. The tasks are designed to help you be sure you know the most important parts of each lesson.

## Course calendar

Tasks are due the Monday at 9:00 Atlantic in the week after they are assigned. Assignment and project component due dates are listed below. Dates for non-synchronous components are guides to help you  pace your work. 

**This is a tentative schedule which will be revised throughout the term.**

```{r echo=FALSE}
my_paste= function(a,b) {
  case_when( is.na(a) | is.na(b) ~ as.character(NA),
             TRUE ~ paste0("[", a, "](", b, ")")
  )
}
row_packing <- table(cal$Week)
attr(row_packing, "names") <- paste0("Week ", attr(row_packing, "names"))
options(knitr.kable.NA = '')
cal %>% select(-Day, -Week) %>%
  arrange(Date) %>%
  mutate(Date = strftime(Date+12*3600, format = "%a %e %b", tz = "America/Halifax"), # timezone error? Excel bug?
         Notes = my_paste(notes, notes_link),
         Video = my_paste(video, video_link),
         Slides = my_paste(slides, slides_link),
         Tasks = my_paste(tasks, tasks_link)
         ) %>%
  select(-notes, -notes_link, -tasks, -tasks_link,
         -video, -video_link, -slides, -slides_link) %>%
  select(Date, Lesson, everything()) %>%
  unite("Links", Notes:Tasks, sep = ", ", na.rm=TRUE) %>%
  kable() %>%
  pack_rows(index = row_packing) %>%
  kable_styling() 
  
```

## Bonus Material

- [Data sources](#data-sources)
- [Staying organized](#organization)
- Getting help: solving problems and asking questions
- [A bit more about R](#r-details). 
- Using R markdown for reproducibility and report writing
- Collaborating with git
- A tour of plot types
- Learning computing tools on your own (project overview)
- What is the tidyverse? Contrast with other conventions
- How do I know I've done it right?
- Text data and regular expressions introduction
- Examples of building up figures step by step
- Working with dates and times

