# Detailed outline {#outline}

```{r include=FALSE}
library(tidyverse)  # fussy, but might be useful later on
library(knitr)
library(kableExtra)
library(readxl)
library(lubridate)
cal <- read_excel("schedule.xlsx")
```

Synchronous sessions will be held on Thursday starting at 8:35 and ending at 9:55 or earlier. You can find the link to these sessions on [Brightspace](https://dal.brightspace.com/d2l/le/content/159870/Home?itemIdentifier=D2L.LE.Content.ContentObject.ModuleCO-2083791). There is an [agenda](#sync) for each session.  Most sessions will be recorded and posted on Brightspace. The other synchronous time in the timetable (Tuesday starting at 8:35) will be used for drop-in office hours. 

Each lesson has written notes, a recorded video, slides, and links to textbook readings in the notes. Some students will prefer to begin with the video, others will prefer to begin with the written notes and slides. I suggest you learn which you prefer by experimentation. The tasks are designed to help you be sure you know the most important parts of each lesson.

## Course calendar

Tasks are due the Monday at 9:00 Atlantic in the week after they are assigned. Assignment and project component due dates are listed below. Dates for non-synchronous components are guides to help you  pace your work. 

**This is a tentative schedule which will be revised throughout the term.**

```{r echo=FALSE}
my_paste= function(a,b) {
  case_when( is.na(a) | is.na(b) ~ as.character(NA),
             TRUE ~ paste0("[", a, "](", b, ")")
  )
}
row_packing <- table(cal$Week)
attr(row_packing, "names") <- paste0("Week ", attr(row_packing, "names"))
options(knitr.kable.NA = '')
cal %>% select(-Day, -Week) %>%
  arrange(Date) %>%
  mutate(Date = strftime(Date+12*3600, format = "%a %e %b", tz = "America/Halifax"), # timezone error? Excel bug?
         Notes = my_paste(notes, notes_link),
         Video = my_paste(video, video_link),
         Slides = my_paste(slides, slides_link),
         Tasks = my_paste(tasks, tasks_link)
         ) %>%
  select(-notes, -notes_link, -tasks, -tasks_link,
         -video, -video_link, -slides, -slides_link) %>%
  select(Date, Lesson, everything()) %>%
  unite("Links", Notes:Tasks, sep = ", ", na.rm=TRUE) %>%
  kable() %>%
  pack_rows(index = row_packing) %>%
  kable_styling() 
  
```

## Additional detail

### Week 3: Jan 18-22

Connect your github account to our course on github and get access to the repository for Task 4 and Assignment 1 by clicking [here](https://classroom.github.com/a/Qb-MKdW3). You will be asked to join the github organization "Dalhousie-AndrewIrwin-Teaching" and then select your name from a list of students. If your name does not appear in the list, please contact me so I can add your name to the list of students. A detailed walk-through of the steps for Task 4 is available as [video](https://dal.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=540e63cc-65b7-441f-b234-aca60134e6bd) and [slides](slides/06-demo-assignment-github.html).

### Week 4: Jan 25-29

- Lesson 11 Summarizing data with dplyr (group_by, filter, mutate, summarize)
- Lesson 12 Format tables
- Synchronous meeting Thursday 28 January, 8:35 Atlantic on Collaborate. Live coding material: [R markdown](static/live-coding-03.rmd), [html](static/live-coding-03.html).  [Agenda](#sync).

Tasks 5 and 6 are due Monday 1 February at 9:00 am Atlantic time. Complete both tasks using the same repository. 

Assignment 2 is due Wednesday 3 February at 6:00 pm Atlantic time.

Click the following links to get two repostories: [tasks-5+6](no link yet) and [assignment-2](no link yet).


### Week 5: Feb 1-5

- [Lesson 13](#reading-data) Reading data (csv, tsv, xlsx); import menu item; r commands
- Office hours. Tuesday 2 February, 8:35 Atlantic on Collaborate.
- [Lesson 14](#pivot) Reshaping data
- [Lesson 15](#facets) Plotting small multiples (facets)
- Task 7 and 8. Check out from and submit on github.

Now review  Show the right numbers. [Healy Chapter 4](https://socviz.co/groupfacettx.html#groupfacettx)
Display and graph tables, add labels, make notes. [Healy Chapter 5](https://socviz.co/workgeoms.html#workgeoms)


Tasks 7 and 8 are due Monday 8 February at 9:00 am Atlantic time. Complete both tasks using the same [repository](no link yet). 

### Week 6: Feb 8-12

- Lesson 16 Work with models. [Healy Chapter 6](https://socviz.co/modeling.html#modeling)
- Lesson 17 Linear models (lines, polynomials, robust, quantile); equationomatic 
- Lesson 18 LOESS, GAM

Tasks 11 and 12 are due Monday 22 February at 9:00 am Atlantic time. Complete both tasks using the same repository. 
Assignment 3 is due Friday 12 February at 6:00 pm Atlantic time. (Change in due date to keep time free during the break for other projects.)

Follow these links to get two repostories: [tasks-11+12](no link yet) and [assignment-3](no link yet).

You will receive an invitation to a term project repository on Friday 12 February.

### Week 7: Feb 22-26

- Lesson 19 PCA
- Lesson 20 MDS 
- Lesson 21 K-means

Tasks 13 and 14 are due Monday 1 March  at 9:00 am Atlantic time. Complete both tasks using the same repository. 

### Week 8: Mar 1-5

- Reproducible reports and R markdown
- Slide presentations using R markdown

Tasks 15 and 16 are due Monday 8 March  at 9:00 am Atlantic time. Complete both tasks using the same repository. 

Assignment 4 is due Wednesday 10 March at 6:00 pm Atlantic time.

### Week 9: Mar 8-12

- Lesson 22 Tidy data; Making data tidy (datapasta)
- Lesson 23 Data validation (pointblank)
- Lesson 24 Finding, describing data with metadata (codebook)

Tasks 17 and 18 are due Monday 15 March  at 9:00 am Atlantic time. Complete both tasks using the same repository. 

- Project proposal due Friday 12 March at 6:00 pm Atlantic.

### Week 10: Mar 15-19

- Lesson 25 Draw maps. [Healy Chapter 7](https://socviz.co/maps.html#maps)
- Lesson 26 Making maps and chloropleths. Heatmaps.
- Lesson 27 Refine your plots. [Healy Chapter 8](https://socviz.co/refineplots.html#refineplots). Ordering factors.

- Synchronous session. Thursday 18 March. Review of selected work from Assigment 4 (sharing ideas and techniques)

Tasks 19 and 20 are due Monday 22 March  at 9:00 am Atlantic time. Complete both tasks using the same repository. 

Assignment 5 is due Wednesday 24 March at 6:00 pm Atlantic time.

### Week 11: Mar 22-26

- Lesson 28 Themes
- Lesson 29 Annotations
- Lesson 30 Graphics output (PDF, png, sizes, colors, themes; ragg) repurposing graphics for different media and purposes

- Synchronous session. Thursday 18 March. Review of selected work from Assigment 5 (sharing ideas and techniques) and problem solving for visualizations you are trying to create for presentation in a week. Send me your questions or challenges in advance.

Tasks 21 and 22 are due Monday 29 March  at 9:00 am Atlantic time. Complete both tasks using the same repository. 

### Week 12: Mar 29-April 1

- graphs for categorical data
- graphs for quantitative data
- Working with dates

- Synchronous session. Thu 1 April, 8:35 Atlantic. Oral presentations day 1. [Agenda](#sync).

No more tasks.

Assignment 6 due Wednesday 7 April at 6:00 pm. Peer assessment of two randomly selected oral presentations. Submit on github.

### Week 13: Apr 5-8

- Synchronous session. Tue 6 April, 8:35 Atlantic. Oral presentations day 2. [Agenda](#sync).

Project report due Friday 9 April, 6:00 pm Atlantic. Submit on github.

## Bonus Material

- [Data sources](#data)
- [Staying organized](#organization)
- Getting help: solving problems and asking questions
- [A bit more about R](#r-details). 
- Using R markdown for reproducibility and report writing
- Collaborating with git
- A tour of plot types
- Learning computing tools on your own (project overview)
- What is the tidyverse? Contrast with other conventions
- How do I know I've done it right?
- Text data and regular expressions introduction
- Examples of building up figures step by step
- Working with dates and times

