# Format tables {#format-tables}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(palmerpenguins)
library(tidyverse)
library(knitr)
library(kableExtra)
```

We're spending a lot of time in this course making graphic visualizations of data. In this one lesson we will take a detour and learn to make nice looking tables with text and numbers in them. There are a bunch of packages for doing this (see Assignment X). We'll use the `kable` table formatting function in this lesson (from the packages `knitr` and `kableExtra`). 

Let's use a simple table to demonstrate the main features. I want some text, some numbers, and not too many rows or columns. I'll summarize the data in the `palmerpenguins` table into a small table.

```{r echo = FALSE}
t1 <- penguins %>% group_by(species, island) %>%
  summarize(count = sum(!is.na(body_mass_g)),
            body_mass_g = mean(body_mass_g, na.rm=TRUE), 
            .groups = "drop")
t1
```

The function `kable` turns this into a formatted table.

```{r}
kable(t1)
```

## Header rows

Often you will want to rename or reformat the column headings. There are two ways to do this: using `rename` to change the actual column names before formatting the table, or using column formatting to just affect the table appearance. You can rename just one column using `rename`, but with the second version below (`col.names`) you need to provide all the column names in the correct order. 

```{r}
t1 %>% rename(`Body mass (g)` = body_mass_g) %>% kable()
```

```{r}
t1 %>% kable(col.names = c("Species", "Island", "Count",  "Body mass (g)"))
```

## Column alignment

Commonly numbers are right justified and text is left justified. That's what's done automatically. You can specify each column as left, centre, or right justified using the letters l, c, or r for each column. Here we'll center the justify the numbers to demonstrate.

```{r}
t1 %>% kable(align = "llcc")
```

## Formatting numbers

Any number that comes from a calculation (such as a mean) will have a lot of decimal places displayed unless you change this. You can control the number of decimal places to show using rounding. (Use a negative number of digits to round to the left of the decimal point, for example `digits=-1` to round to the tens place.) Give either one number to use for all columns, or provide a vector to control the number of digits separately for each column.

You can add a comma (or space for SI or . for European styles) to separate the thousands or millions using `format.args = list(big.mark = ",")`. See the help for `format` for more options.


```{r}
t1 %>% kable(digits = 1, format.args = list(big.mark = ","))
```

## Color, highlights, and other styles

There are a lot of options for changing the appearance of text in the `kableExtra` package. If you are interested, look at the [vignette](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html) linked in the further reading.

The two styles I use frequently are alternating shading to help you read across rows and making the columns only wide enough to display your data.

```{r}
t1 %>% kable() %>% kable_styling(full_width = FALSE, bootstrap_options = "striped")
```


## Captions

You can add a caption to a table with the `caption` argument to `kable`. 

```{r}
t1 %>% kable(caption = "The number and average mass of penguins by species and island.")
```


## Putting it all together

Usually you will want to combine these features to get the look you want. Your goal should always be to make a table that clearly displays your data.

```{r}
t1 %>% kable(digits = 1, format.args = list(big.mark = ","),
             col.names = c("Species", "Island", "n", "Body mass (g)"),
             caption="The number and average mass of penguins by species and island.") %>%
  kable_styling(full_width = FALSE, bootstrap_options = "striped")
```

## Further reading

* Using [kable](https://bookdown.org/yihui/rmarkdown-cookbook/kable.html) to format tables
* Vignette on using [kable and kableExtra](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html)
* https://people.ok.ubc.ca/jpither/modules/Tables_markdown.html

