---
title: "Data Visualization"
subtitle: "Making your first plot"
author: "Andrew Irwin, a.irwin@dal.ca"
institute: "Math & Stats, Dalhousie University"
date: "2021-01-13 (updated: `r Sys.Date()`)"
ratio: "16:9"
output:
  xaringan::moon_reader:
    css: [default, hygge, useR-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
library(patchwork)
library(kableExtra)
```

# Use the ggplot2 library

Do this once (if you haven't done it already):

```
install.packages("tidyverse")
```

Add this line to every R markdown document:

```{r}
library(tidyverse)
```

---
class: middle

# Get some data

```{r}
str(mtcars)
```

---

# First plot

```{r fig0, fig.show="hide", warning=FALSE}
mtcars %>%
  ggplot(aes(x = wt,
             y = mpg)) +
  geom_point()
```
--
.pull-left[
The "pipe" symbol (`%>%`) is function composition. `f(g(x))` can be written `x %>% g %>% f`.

`aes` is a function to define aesthetic associations between features of your plot and variables in the dataset.

Parts of a ggplot are added togther with `+`

The kind of plot is called its "geometry". A scatterplot is `geom_point`.
]

--
.pull-right[
```{r fig.align="center", out.width="100%", echo=FALSE}
knitr::include_graphics(knitr::fig_chunk("fig0", "png"))
```
]


---
# What if you forget ... ?

```{r fig1, fig.show="hide", warning=FALSE}
mtcars %>%
  ggplot(aes(x = wt,
             y = mpg))
```

--
.pull-left[
Try "forgetting" other parts of the code to see what goes wrong.
]
.pull-right[
```{r fig.align="center", out.height="80%", echo=FALSE}
knitr::include_graphics(knitr::fig_chunk("fig1", "png"))
```
]


---
# Add some colour

```{r fig2, fig.show="hide", warning=FALSE, fig.height=6, fig.width=8}
mtcars %>%
  ggplot(aes(x = wt,
             y = mpg,
             color = factor(cyl))) + #<<
  geom_point()
```

--
.pull-left[
`cyl` is a number, so I must turn it into a categorical variable (factor) to get a discrete colour scale.

Try using `cyl` instead of `factor(cyl)`.
]
.pull-right[
```{r fig.align="center", out.height="80%", echo=FALSE}
knitr::include_graphics(knitr::fig_chunk("fig2", "png"))
```
]

---
# Make the text larger

```{r fig3, fig.show="hide", warning=FALSE, fig.width=8, fig.height=6}
mtcars %>%
  ggplot(aes(x = wt,
             y = mpg,
             color = factor(cyl))) + 
  geom_point() +
  theme(text = element_text(size = 28)) #<<
```

---
# Make the text larger

```{r fig.align="center", out.width="80%", out.height="80%", echo=FALSE}
knitr::include_graphics(knitr::fig_chunk("fig3", "png"))
```


---
# Make the symbols larger

```{r fig3b, fig.show="hide", warning=FALSE, fig.width=8, fig.height=6}
mtcars %>%
  ggplot(aes(x = wt,
             y = mpg,
             color = factor(cyl))) + 
  geom_point(size=3) + #<<
  theme(text = element_text(size = 28)) 
```

--

I'm setting the size of all the symbols, not connecting a variable to the size aesthetic. So don't use `aes`.

---
# Make the symbols larger

```{r fig.align="center", out.width="80%", out.height="80%", echo=FALSE}
knitr::include_graphics(knitr::fig_chunk("fig3b", "png"))
```


---
# Customize the labels

```{r fig4, fig.show="hide", warning=FALSE, fig.width=8, fig.height=6}
mtcars %>%
  ggplot(aes(x = wt,
             y = mpg,
             color = factor(cyl))) + 
  geom_point(size=3) +
  theme(text = element_text(size = 28)) +
  labs(x = "Car mass (x 1000 pounds)", #<<
       y = "Fuel consumption (mpg)", #<<
       color = "Number of\nCylinders", #<<
       title = "Car road test results", #<<
       caption = "from Motor Trend magazine") #<<
```

---
# Customize the labels

```{r fig.align="center", out.width="80%", out.height="80%", echo=FALSE}
knitr::include_graphics(knitr::fig_chunk("fig4", "png"))
```

---
class: middle

# Summary

* Start with data

* Pipe (`%>%`) into `ggplot`

* Define the aesthetics: `aes(x = ..., y = ..., color = ..., shape = ...)`

* Define the geometry: `geom_point` shown here, but there are lots more

* Customize text

---
class: middle

# Suggested reading

* Course notes: Making your first plot

* Healy. Section 2.6. Make your first figure

* R2DS. Chapter 3: Data visualization

---
class: middle, inverse

# Task and Assignment

* Try these plotting commands on your own 

* Assigment 1: Your first plotting exercises

  * Do Task 4 following Lesson 6 on git and github first


---
# Datasets to experiment with

* mtcars, iris and many other well-known data in datasets package
* penguins in palmerpenguins package
* gapminder in gapminder package  (but see website too [Gapminder](https://gapminder.org))
* diamonds in ggplot2 package
* nycflights13 in dbplyr package

Use `str(mtcars)` (or other dataset) and `View(mtcars)` to look at the data.

Use ?mtcars to get documentation for the dataset, or use the search in the "Help" panel.

