---
title: "Data Visualization"
subtitle: "Splitting graphs into facets"
author: "Andrew Irwin, a.irwin@dal.ca"
institute: "Math & Stats, Dalhousie University"
date: "2021-01-27 (updated: `r Sys.Date()`)"
ratio: "16:9"
output:
  xaringan::moon_reader:
    css: [default, hygge, useR-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
library(palmerpenguins)
library(gapminder)
```

# Splitting a plot into facets

* Using one categorical variable

* Using two categorical variables

* Using quantitative variables

---
class: middle, inverse
# One categorical variable

* `facet_wrap( ~ variable)`

---
class: middle

```{r}
mtcars %>% ggplot(aes(x =wt, y=mpg)) + geom_point()
```

---
class: middle

```{r}
mtcars %>% ggplot(aes(x =wt, y=mpg)) + geom_point() +
  facet_wrap(~cyl)
```

---
class: middle

```{r warning=FALSE}
penguins %>%
  ggplot(aes(x=body_mass_g, y= flipper_length_mm, color = sex)) +
  geom_point() + 
  facet_wrap( ~ species)
```


---
class:middle, inverse

# Using two categorical variables

* `facet_grid( y ~ x)`
* `facet_wrap( ~ x + y)`
---
class: middle

```{r}
mtcars %>% ggplot(aes(x =wt, y=mpg)) + geom_point() +
  facet_grid(gear~cyl)
```

---
class: middle

```{r}
mtcars %>% ggplot(aes(x =wt, y=mpg)) + geom_point() +
  facet_grid(gear~cyl, 
             labeller = labeller(.cols = function(x) paste(x, " cylinders"),
                                 .rows = label_both))
```


---
class: middle

```{r warning=FALSE}
penguins %>%
  ggplot(aes(x=body_mass_g, y= flipper_length_mm, color = sex)) +
  geom_point() + 
  facet_wrap(~ species + island)
```


---
class: middle

```{r}
gapminder %>%
  ggplot(aes(x=year, y = lifeExp, group = country)) + 
  geom_line() +
  facet_grid( continent ~ . )
```

---
class: middle

```{r message=FALSE}
gapminder %>%
  ggplot(aes(x=year, y = lifeExp)) + 
  geom_line(aes(group = country), color = "gray40", size=0.25) +
  geom_smooth(method = "lm", color = "blue", se=FALSE) + 
  facet_grid( continent ~ . )
```

---
class: middle, inverse

# Splitting on a quantitative variable

* Start by converting the quantitative variable into a cateogorical one

For example:

```{r}
penguins %>% 
  mutate(bill_categories = cut_number(bill_length_mm, n = 4)) %>%
  select(species, bill_length_mm, bill_categories, everything())
```

---
class: middle

```{r warning=FALSE}
penguins %>% 
  mutate(bill_categories = cut_number(bill_length_mm, n = 3)) %>%
  ggplot(aes(x= body_mass_g, y = flipper_length_mm, color = species)) + 
  geom_point() + 
  facet_wrap( ~ bill_categories)
```

---
class: middle

# Summary

* `facet_wrap()`, `facet_grid()`

* You can set the number of rows and columns with `nrow = Y` or `ncol=X`

* You can label facets with both the variable and value using `labeller = labeller(.rows = label_both)`


---
class: middle

# Further reading

* More examples with different data in the course notes

* R4DS [Section 3.5](https://r4ds.had.co.nz/data-visualisation.html#facets)

* Wilke [Multi-panel figures](https://clauswilke.com/dataviz/multi-panel-figures.html) discussing facets and other suggestions for combining different views of data together into one figure

* Healy [Section 4.3](https://socviz.co/groupfacettx.html#facet-to-make-small-multiples)

---
class: middle, inverse

# Task

Practice these skills by making the plots in Task 6.

