---
title: "Data Visualization"
subtitle: "Summarizing data (dplyr)"
author: "Andrew Irwin, a.irwin@dal.ca"
institute: "Math & Stats, Dalhousie University"
date: "2021-01-25 (updated: `r Sys.Date()`)"
ratio: "16:9"
output:
  xaringan::moon_reader:
    css: [default, hygge, useR-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
library(rmarkdown)
library(knitr)
library(kableExtra)
```

# Transforming data

How can we  
  * transform data, or 
  * reduce a large number of observations to a smaller number?

* Compute derived quanties (ratios, change units) 

* Count observations in categories

* Compute means and standard deviations 
  * Combine replicates
  * Average over cateogories

* Highlight special cases

---
class: middle, inverse
# dplyr examples

Transforming and reducing observations using 

* `mutate`

* `filter`

* `group_by`

* `summarize`

We will use the `ChickWeight` data in the `datasets` package.


---
class: middle

# Data

```{r}
glimpse(ChickWeight)
```

---
class: middle
# Use summarize to count observations

```{r}
ChickWeight %>% summarize(count = n()) %>% kable()
```

---
class: middle
# Count observations in groups

```{r}
ChickWeight %>%
  group_by(Diet) %>% #<<
  summarize(count = n(), .groups="drop") %>%
  kable()
```


---
class: middle
## Two grouping factors

```{r}
ChickWeight %>%
  group_by(Diet, Chick) %>% #<<
  summarize(count = n(), .groups = "drop") %>%
  paged_table()
```

---
class: middle
### Compute any function you want

```{r}
ChickWeight %>%
  group_by(Chick, Diet) %>% 
  summarize(count = n(), 
            max_time = max(Time), 
            median_weight = median(weight),
            increase_weight = max(weight) - min(weight),
            .groups="drop") %>%
    paged_table()
```

---
class: middle
### Use mutate to create new columns for each observation

```{r}
ChickWeight %>% as_tibble() %>%
  mutate(growth = weight / Time) %>%
  paged_table()
```

---
class: middle
### Use mutate to create new columns for each observation

```{r}
ChickWeight %>% as_tibble() %>%
  mutate(growth = (weight-min(weight)) / Time) %>%
  paged_table()
```

---
class: middle
### Exclude observations with filter

```{r}
ChickWeight %>% as_tibble() %>%
  filter(Time >= 10) %>% 
  arrange(Time) %>%
  paged_table()
```


---
class: middle
### Exclude observations with filter

```{r}
ChickWeight %>% as_tibble() %>%
  filter(Time == 10) %>% 
  group_by(Diet) %>% 
  summarize(weight_10 = mean(weight), .groups = "drop") %>%
  kable()
```

---
class: middle
# Combine all three

* Compute max weight for each chick and diet over all time.
* Filter out chicks that reach a target weight
* Count the number of observations per chick and diet treatment

```{r eval=FALSE}
ChickWeight %>% as_tibble() %>%
  group_by(Diet, Chick) %>%
  mutate(max_weight = max(weight)) %>%
  filter(max_weight > 110) %>%
  group_by(Diet, Chick) %>%
  summarize(n = n()) %>%
  summarize(count_chicks = n(), .groups = "drop") %>%
  kable()
```

---
class: middle
# Combine all three

* Compute max weight for each chick and diet over all time.
* Filter out chicks that reach a target weight
* Count the number of observations per chick and diet treatment

```{r echo=FALSE, message=FALSE}
ChickWeight %>% as_tibble() %>%
  group_by(Diet, Chick) %>%
  mutate(max_weight = max(weight)) %>%
  filter(max_weight > 110) %>%
  group_by(Diet, Chick) %>%
  summarize(n = n()) %>%
  summarize(count_chicks = n(), .groups = "drop") %>%
  kable()
```

---
class: middle

# Summary

* Use `mutate` to define a new variable for all observations

* Use `summarize` to compute functions (count, mean, ...) to reduce many observations to one

* Use `group_by` to split, summarize, and combine data

* Use `filter` to select observations


---
class: middle

# Further reading

* Lots more examples with different data in the course notes

* R4DS [Chapter 5 Data transformations](https://r4ds.had.co.nz/transform.html)

---
class: middle, inverse

# Task

Practice these skills by doing the calculations in Task 5.

