---
title: "Data Visualization"
subtitle: "Reshaping data"
author: "Andrew Irwin, a.irwin@dal.ca"
institute: "Math & Stats, Dalhousie University"
date: "2021-02-01 (updated: `r Sys.Date()`)"
ratio: "16:9"
output:
  xaringan::moon_reader:
    css: [default, hygge, useR-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
library(kableExtra)
```

# Wide and Long data

* What is tidy data?

* What is the difference between wide and long data?

* Why would you want data in long format?

* Why would you want data in wide format?

* Pivoting data

---
class: middle, inverse
# Tidy data

* One observation per row

* Each column is a different variable

* There are many ways to organize data. A key idea behind tidy data is to standardize the way data are organized, to make working with data simpler

---
class: middle
## Example 1 (tidy)

```{r echo=FALSE}
df1 <- tribble(
  ~country,~year, ~cases, ~population,
"Afghanistan",  1999,    745,  19987071,
"Afghanistan",  2000,   2666,   20595360,
"Brazil",       1999,  37737,  172006362,
"Brazil",       2000,  80488,  174504898,
"China",        1999, 212258, 1272915272,
"China",        2000, 213766, 1280428583,
)
df1 %>% kable() %>% kable_styling(full_width = FALSE)
```

---
class: middle
## Example 2 (long)

```{r echo=FALSE}
df1 %>% pivot_longer(cases:population, names_to = "type", values_to = "count") %>%
  kable() %>% kable_styling(full_width = FALSE)
```

---
class: middle
## Example 3

```{r echo=FALSE}
df1 %>% unite("rate", cases:population, sep="/") %>%
  kable() %>% kable_styling(full_width = FALSE)
```

---
class: middle
## Example 4

```{r echo=FALSE}
df1 %>% select(-population) %>%
  pivot_wider(names_from = "year", values_from = "cases") %>%
  kable(caption="Cases") %>% kable_styling(full_width = FALSE)
```

```{r echo=FALSE}
df1 %>% select(-cases) %>%
  pivot_wider(names_from = "year", values_from = "population") %>%
  kable(caption="Population") %>% kable_styling(full_width = FALSE)
```


---
class: middle
## Where do data organization patterns come from?

* The way data are collected (population from one database, cases from another)

* The way data are recorded (one year at a time)

* Plans for how data will be interpreted or stored

* Concerns about readability (any data missing?) or compactness (storage, display)

* Habits


---
class: middle
## What is good about tidy data?

* Easy to use `dplyr`: filter, group_by, summarize

* Easy to use `ggplot`: aesthetic mappings (x, y, color, shape) and facets

---
class: middle, inverse
## Pivoting

### Step by step examples of the transformations shown above

---
class: middle
## Pivot longer

.left-column[.tiny[
```{r echo=FALSE}
df1 %>%
  kable() %>% kable_styling(full_width = FALSE)
```
]]

.right-column[.small[
```{r}
df1 %>% 
  pivot_longer(cases:population, 
               names_to = "type", 
               values_to = "count") %>%
  kable() %>% kable_styling(full_width = FALSE)
```
]]

---
class: middle
## Unite, mutate

```{r}
df1 %>% mutate(rate_fraction = cases/population) %>%
  unite("rate", cases:population, sep="/") %>%
  kable() %>% kable_styling(full_width = FALSE)
```

---
class: middle
## Separate

```{r}
df1 %>% unite("rate", cases:population, sep="/") %>%
  separate(rate, into = c("cases", "population"), sep="/") %>%
  kable() %>% kable_styling(full_width = FALSE)
```

---
class: middle
## Make two tables

```{r}
df1A <- df1 %>% select(-population) %>%
  pivot_wider(names_from = "year", values_from = "cases") 
df1A %>% kable(caption="Cases") %>% kable_styling(full_width = FALSE)
```

```{r echo=FALSE}
df1B <- df1 %>% select(-cases) %>%
  pivot_wider(names_from = "year", values_from = "population") 
df1B %>% kable(caption="Population") %>% kable_styling(full_width = FALSE)
```

---
class: middle
## Combine two tables into one

```{r}
bind_rows(bind_cols(type = "cases", df1A),
          bind_cols(type = "population", df1B)) %>%
  kable() %>% kable_styling(full_width = FALSE)
```

---
class: middle
# Summary

* What and why of tidy data, longer vs wider format

* Common reshaping tasks


---
class: middle

# Further reading

* Tidy Data (Chapter 12) in R4DS

---
class: middle, inverse

## Task

Practice these reshaping skills in Task 7

