---
title: "Data Visualization"
subtitle: "Displaying tables"
author: "Andrew Irwin, a.irwin@dal.ca"
institute: "Math & Stats, Dalhousie University"
date: "2021-02-03 (updated: `r Sys.Date()`)"
ratio: "16:9"
output:
  xaringan::moon_reader:
    css: [default, hygge, useR-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
library(kableExtra)
```

# Displaying tables

* After we collect, reshape, and summarize data, we want to display it!

* Rename and align columns
* Format (bold, italic, colour)
* Add column headings
* Group rows
* Add captions

* Consistent goal: make the table easy to interpret

---
class: middle

## Raw data

.tiny[
```{r echo = FALSE}
mtcars
```
]



---
class: middle
## `kable` formatting

```{r warning=FALSE, message=FALSE, fig.height=4.25, fig.align="center", dpi=150}
mtcars %>% kable() %>%
  kable_styling(full_width = FALSE, 
                bootstrap_options = c("striped", "condensed"))
```


---
class: middle
## Scrolling to show big tables

```{r}
mtcars %>% kable() %>% 
  kable_styling() %>% scroll_box()
```

---
class: middle
## Paged data to show big tables

```{r}
mtcars %>% rmarkdown::paged_table()
```

---
class: middle
## Changing column headings

```{r}
mtcars %>% as_tibble(rownames = "make_model") %>%
  select(make_model, mpg, cyl, wt) %>%
  kable(col.names = c("Make & Model", "Miles per gallon", 
                      "Cylinders", "Weight (1000 lbs)")) %>%
  kable_styling(full_width = FALSE, bootstrap_options = "striped")
```


---
class: middle
## Formatting rows & columns

```{r}
mtcars %>% as_tibble(rownames = "make_model") %>% select(make_model, mpg, cyl, wt) %>%
  kable(col.names = c("Make & Model", "Miles per gallon", "Cylinders", "Weight (1000 lbs)"),
        align = "lccr",
        digits = 1) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(3, color = "blue") %>%
  row_spec(2, italic = TRUE) %>%
  kable_styling(full_width = FALSE, 
                font_size=24,
                bootstrap_options = "striped")
```


---
class: middle
## Headers spanning multiple columns

```{r}
mtcars %>% select(mpg, cyl, hp, wt) %>% kable() %>%
  add_header_above(c(" " = 1, "Group A" = 2, "Group B" = 2)) %>%
  kable_styling(full_width = FALSE, bootstrap_options = "striped")
```


---
class: middle
## Grouping rows

```{r}
mtcars %>% select(mpg, cyl, hp, wt) %>% kable() %>%
  group_rows("Group 1", 1, 4) %>%
  group_rows("Group 2", 5, 9) %>%
  kable_styling(full_width = FALSE, bootstrap_options = "striped")
```


---
class: middle
## Adding caption, footnote

```{r}
mtcars %>% head() %>% select(mpg, cyl, hp, wt) %>% 
  kable(caption="Car data from Motor Trend magazine.") %>%
  kable_styling(full_width = FALSE, bootstrap_options = "striped") %>%
  footnote(symbol = c("Footnote 1.", 
                      "Footnote 2")) 
```


---
class: middle

# Summary

* Several ways to customize the formatting of a table



---
class: middle

# Further reading

* Course notes

* `kableExtra` vignette


---
class: middle, inverse

## Task

Task 8 has some exercises for practing table presentation.



