---
title: "Data Visualization"
subtitle: "Getting help"
author: "Andrew Irwin, a.irwin@dal.ca"
institute: "Math & Stats, Dalhousie University"
date: "2021-02-04 (updated: `r Sys.Date()`)"
ratio: "16:9"
output:
  xaringan::moon_reader:
    css: [default, hygge, useR-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
library(gapminder)
```

## What to do when you are stuck?

* How do I ... ?

* Why did R do X when I wanted it to do Y?

* My code doesn't work!

* Help!!

---
class: middle, inverse

## Learning technical skills is challenging

* There are many ways to approach a task

* You need to figure out how to describe what you want to accomplish and how to do it

* You need to evaluate your decisions -- is that hard-won figure effective?

* You need a strategy for making notes and building your own knowledge

* Even small errors can be hard to find and cause a lot of frustration

---
class: middle
## Ways to get help

* Finding and fixing mistakes

* Learning from examples

* Remembering what you learned before

* Learning something new

* Asking questions

---
## Finding and fixing mistakes

```{r}
gapminder %>%
  filter(country == "Canada", country == "Iran")
```

--

```{r}
gapminder %>%
  filter(country == "Canada" | country == "Iran")  # logical "or"
```


---
## Finding and fixing mistakes

```{r error=TRUE}
gapminder %>%
  filter(country == "Canada" | country == "Iran") %>%
  ggplot(aes(x = Year, y = LifeExp, color = Country)) +
  geom_point()
```

---

```{r}
gapminder %>%
  filter(country == "Canada" | country == "Iran") %>%
  ggplot(aes(x = year, y = lifeExp, color = country)) +
  geom_point()
```

---
class: middle
# Learning from examples

Look in the **Examples** section of `?geom_line`

```{r fig.height=4}
ggplot(economics, aes(unemploy, date)) + 
  geom_line(orientation = "y")  #<<
```

---
class: middle
# Learning from examples

.pull-left[
```{r}
ggplot(economics, aes(unemploy, date)) + 
  geom_line(orientation = "y")  
```
]
.pull-right[
```{r}
ggplot(economics, aes(unemploy, date)) + 
  geom_line()  #<<
```
]


---
class: middle
# Remembering what you learned before

How do I change the x and y axis labels?

* Google search. Add "R" to your search.

* Know which sites you like to read

* Learn to read documentation -- more condensed information
  * https://ggplot2.tidyverse.org/ especially "cheatsheet" and "reference"
  * Scales > Axis > Labs: https://ggplot2.tidyverse.org/reference/labs.html

* Make notes for yourself

---
class: middle
# Learn something new

How do I set the limits, tick marks, and numbering on an axis?

* ggplot2 tidyverse "Scales": `xlim`, `ylim`

* Search. 
  * Find `theme`
  * Find `scale_x_discrete(name, breaks, labels, limits)`
  * Also `scale_x_continuous(name, breaks, lables, limits, trans)`
  
* Try an example from the help page (or internet)

---
class: middle

```{r warning=FALSE}
p1 <- ggplot(mpg, aes(displ, hwy)) + geom_point()
p1 + scale_x_continuous(name = "Displacement", 
                        limits = c(2, 6),
                        breaks = c(2, 4, 6),
                        labels = c("Two", "Four", "Six"))
```


Read a bit of `?c` (combine values into a vector or list).


---
class: middle

# Asking questions

You are still stuck.

How do you ask for help?

* What are you trying to do?
* What have you tried?
* What was the result (output, graphic, error message)?
* What do you think the problem might be?



---
class: middle

# Summary

* Learning computing skills is a never-ending task

* Using reference material, blogs, and books is a learned skill that you can get better at

* Try different sources until you learn what you like to learn from the most (and sometimes try other sources)

* Asking questions is a valuable skill. Clearly expressing your question so that someone else can read it will often help you solve the problem by yourself.

* It gets easier with practice: posing questions, understanding the answers, and remembering how it all fits together


---
class: middle

# Further reading

* Course notes

* Two or more sources referenced in earlier lessons

---
class: middle, inverse

## Task

Assignment 2 asks you to learn some new skills and describe what you learned.



