---
title: "Data Visualization"
subtitle: "Working with models"
author: "Andrew Irwin, a.irwin@dal.ca"
institute: "Math & Stats, Dalhousie University"
date: "2021-02-08 (updated: `r Sys.Date()`)"
ratio: "16:9"
output:
  xaringan::moon_reader:
    css: [default, hygge, useR-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
my_theme = theme_classic() + 
  theme(text = element_text(size=18))
```

## Models and visualizations

* Adding regression and smoothing lines to ggplot visualizations

* Describing, fitting, and using models

  * Linear regression
  
  * Other regression
  
  * Generalized additive models
  
  * LOESS smooths

---
class: middle, inverse

##  Examples

* Linear regression

* Polynomial model

* Quantile regression (fitting line to medians or other quantiles)

* Generalized additive model (piecewise splines)

* LOESS (local linear and quadratic regression)

* Putting multiple smooths on one plot

* Smooths on facetted plots

---
class: middle
###  Start with a basic graph

```{r, fig.height=4.5, fig.align="center"}
mpg %>% mutate(city_l100km = 235.214583 / cty) %>%
   ggplot(aes(x =displ, y= city_l100km)) + geom_point() +
   labs(x = "Engine displacement (L)",
        y = "Fuel economy, city (L/100km)") + my_theme
```
```{r include = FALSE}
p <- mpg %>% mutate(city_l100km = 235.214583 / cty) %>%
       ggplot(aes(x =displ, y= city_l100km)) + geom_point() +
       labs(x = "Engine displacement (L)",
            y = "Fuel economy, city (L/100km)") +
       my_theme
```

---
class: middle
### Add a regression line

```{r, fig.height=5, fig.align="center"}
p + geom_smooth(method = "lm", formula = y ~ x)
```

---
class: middle
### Add a regression line for each group

```{r, fig.height=5, fig.align="center"}
p + geom_smooth(aes(color = factor(cyl)), method = "lm", formula = y ~ x)
```


---
class: middle
###  Polynomial regression 

```{r, fig.height=5, fig.align="center"}
p + geom_smooth(method = "lm", formula = y ~ poly(x,2))
```

---
class: middle
### Quantile regression

```{r, fig.height=5, fig.align="center"}
p + geom_quantile(method = "rqss", formula = y ~ x,
                  quantiles = c(0.1, 0.5, 0.9))
```

---
class: middle
### Generalized additive model 

```{r, fig.height=5, fig.align="center"}
p + geom_smooth(method = "lm", 
                formula = y ~ splines::bs(x, df = 5))
```

---
class: middle
### Generalized additive model 

```{r, fig.height=5, fig.align="center"}
p + geom_smooth(method = "gam", 
                formula = y ~ s(x, bs="cs"))
```

---
class: middle
### LOESS

```{r, fig.height=5, fig.align="center"}
p + geom_smooth(method = "loess", 
                formula = y ~ x) 
```

---
class: middle
### LOESS

```{r, fig.height=5, fig.align="center"}
p + geom_smooth(method = "loess", formula = y ~ x,
                span = 0.3) 
```

---
class: middle
### Putting multiple smooths on one plot

```{r, fig.height=5, fig.align="center", message=FALSE}
p + geom_smooth(aes(color = "Linear"), method = "lm", se=FALSE) +
    geom_smooth(aes(color = "GAM"), method = "gam", se=FALSE) +
    geom_smooth(aes(color = "LOESS"), method = "loess", formula = y ~ x, se=FALSE) +
    labs(color = "Type")
```

---
class: middle
### Smooths on facetted plots

```{r, fig.height=5, fig.align="center"}
p + geom_smooth(method = "lm", se=FALSE, formula = y ~ x) +
  facet_wrap( ~ class)
```

---
class: middle
### Other customizations

```{r, fig.height=5, fig.align="center", message=FALSE}
p + geom_smooth(method = "lm", se=TRUE, 
                fullrange = TRUE, level = 0.50) + xlim(1,8)
```
---
class: middle
### Modelling x as a function of y

```{r, fig.height=5, fig.align="center"}
p + geom_smooth(method = "loess", formula = y ~ x, orientation = "y")
```

---
class: middle
### Modelling x as a function of y

```{r, fig.height=5, fig.align="center"}
p + geom_smooth(method = "lm", formula = y ~ x, orientation = "y") +
  geom_smooth(method = "lm", formula = y ~ x, color = "red")
```

---
class: middle

# Summary

* Regression lines (based on a model) and "smooths" (no simple model) can be used to draw attention to patterns in data

* We are using smooths for their visual effect

* Logical reasoning for the lines you draw is important (next lessons)

---
class: middle

# Further reading

* Course notes

* Healy Chapter 6. Work with models.

---
class: middle, inverse

## Task

Bonus task for this lesson to practice these functions. Nothing to hand in.




