---
title: "Data Visualization"
subtitle: "Linear models"
author: "Andrew Irwin, a.irwin@dal.ca"
institute: "Math & Stats, Dalhousie University"
date: "2021-02-10 (updated: `r Sys.Date()`)"
ratio: "16:9"
output:
  xaringan::moon_reader:
    css: [default, hygge, useR-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
library(broom)
library(equatiomatic)
library(kableExtra)
library(gapminder)
my_theme = theme_classic() + 
  theme(text = element_text(size=18))
```

##  Linear models

* Describing, fitting, and using models

* Linear regression

  * Straight lines
  * Polynomials
  * log transforms
  
* Quantile regression

* Accessing model coefficients 

* Calculating residuals

* Making predictions: point estimates and standard errors or confidence intervals



---
class: middle
###  Start with a basic graph

```{r, fig.height=4.5, fig.align="center", echo = FALSE}
my_mpg <- mpg %>% mutate(city_l100km = 235.214583 / cty) 
p <- my_mpg %>%
   ggplot(aes(x =displ, y= city_l100km)) + geom_point() +
   labs(x = "Engine displacement (L)",
        y = "Fuel economy, city (L/100km)") + my_theme
p
```

---
class: middle
### Fit a straight line

```{r, fig.height=5, fig.align="center"}
m1 <- lm( city_l100km ~ displ, data = my_mpg)
summary(m1)
```


---
class: middle
### Fit a straight line

```{r}
library(broom)
glance(m1)  %>% kable(digits = 2) %>% kable_styling(full_width = FALSE)
tidy(m1) %>% kable(digits = 2) %>% kable_styling(full_width = FALSE)
```

---
class: middle
### Graph estimated terms

```{r fig.height=5, fig.align="center"}
tidy(m1, conf.int = TRUE) %>%
  ggplot(aes(y = term, x = estimate, xmin = conf.low, xmax = conf.high)) + 
  geom_pointrange() + my_theme
```

---
class: middle
### Polynomial regression

```{r, fig.height=5, fig.align="center"}
m2 <- lm( city_l100km ~ poly(displ,2), data = my_mpg)
glance(m2) %>% kable(digits = 2) %>% kable_styling(full_width = FALSE)
tidy(m2) %>% kable(digits = 2) %>% kable_styling(full_width = FALSE)
```


---
class: middle
###  Log transforms

```{r, fig.height=5, fig.align="center"}
gapminder %>% ggplot(aes(x = gdpPercap, y= lifeExp)) + geom_point() +
  scale_x_log10() + 
   geom_smooth(method = "lm", formula = y ~ x) + my_theme
```


---
class: middle
###  Log transforms

```{r, fig.height=5, fig.align="center"}
m3 <- gapminder %>% mutate(logGDPpercap = log10(gdpPercap)) %>%
  lm( lifeExp ~ logGDPpercap, data = . ) 
tidy(m3) %>% kable(digits = 2) %>% kable_styling(full_width = FALSE)
glance(m3) %>% kable(digits = 2) %>% kable_styling(full_width = FALSE)
```

---
class: middle
### Quantile regression

```{r  message=FALSE}
library(quantreg)
m4 <- my_mpg %>% rq(city_l100km ~ displ, data = .,
                      tau = 0.50) # quantile, between 0 and 1
tidy(m4) %>% kable(digits = 2) 
glance(m4) %>% kable(digits = 2) 
```

---
class: middle
### Working with models: predictions

```{r, fig.height=5, fig.align="center"}
predict(m1, se.fit = TRUE) %>% as_tibble %>% kable(digits = 2)
```

---
class: middle
### Working with models: residuals

```{r, fig.height=5, fig.align="center"}
residuals(m1) %>% as_tibble %>% kable(digits = 2)
```

---
class: middle
### Combining with original data

```{r, fig.height=5, fig.align="center"}
bind_cols(my_mpg %>% select(displ, city_l100km),
          predict(m1) %>% as_tibble %>% rename(predict = value),
          residuals(m1) %>% as_tibble %>% rename(residual = value)) %>% kable(digits = 2)
```


---
class: middle
### Predictions from new data

```{r, fig.height=5, fig.align="center"}
new_data = tibble(displ = seq(1.6, 7.2, by = 0.5))
new_data %>% mutate(prediction = predict(m1, new_data)) %>% kable(digits = 2)
```

---
class: middle
### Model uncertainties

```{r, fig.height=5, fig.align="center"}
bind_cols(my_mpg %>%  select(displ, city_l100km),
          predict(m1, se.fit = TRUE) %>% as_tibble) %>%
 distinct() %>% kable(digits = 2)
```

---
class: middle
### Model confidence intervals

```{r, fig.height=5, fig.align="center"}
bind_cols(my_mpg %>% select(displ, city_l100km),
          predict(m1, interval = "confidence") %>% as_tibble) %>%
  distinct() %>% kable(digits = 2)
```


---
class: middle
### Prediction uncertainties

```{r, fig.height=5, fig.align="center"}
bind_cols(new_data,
          predict(m1, new_data, interval = "prediction") %>% as_tibble) %>%
   kable(digits = 2)
```

---
class: middle
### Can also use augment

```{r}
augment(m1, my_mpg, interval = "confidence") %>%
  select(displ, city_l100km, .fitted, .lower, .upper, .resid, everything() ) %>% kable(digits =2)
```

---
class: middle
### Same methods for other model types

```{r}
augment(m2, newdata = new_data, interval = "prediction") %>%
  kable(digits = 2)
```


---
class: middle
### Same methods for other model types

```{r}
new_data2 = tibble(logGDPpercap = seq(2.3, 5.0, 0.25))
augment(m3, newdata = new_data2, 
        interval = "prediction", level = 0.90) %>%
   kable(digits = 2)
```

---
class: middle
### Same methods for other model types

```{r}
augment(m4, newdata = new_data, interval = "confidence") %>% kable(digits = 2)
```

---
class: middle

# Summary

* Fitting regression models (lines, polynomials, transformations) to means and quantiles

* Extracting model coefficients, uncertainties, errors (with `tidy`) 

* Computing predicted values (with two kinds of errors) and residuals (compared to data)

* Most models can't make predictions from NA. NAs in data can cause problems.

---
class: middle

# Further reading

* Course notes

* Healy Chapter 6. Work with models.

---
class: middle, inverse

## Task

Task 9: Fitting and using linear models.





