---
title: "Data Visualization"
subtitle: "GAM and LOESS models"
author: "Andrew Irwin, a.irwin@dal.ca"
institute: "Math & Stats, Dalhousie University"
date: "2021-02-12 (updated: `r Sys.Date()`)"
ratio: "16:9"
output:
  xaringan::moon_reader:
    css: [default, hygge, useR-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
library(mgcv)
library(broom)
library(gratia)
library(kableExtra)
library(lubridate)
my_theme = theme_classic() + 
  theme(text = element_text(size=18))
```

##  GAM and LOESS models

* Describing, fitting, and using models

* No model coefficients 

* Calculating residuals

* Making predictions: point estimates and standard errors or confidence intervals



---
class: middle
###  Start with a basic graph

```{r, fig.height=4.5, fig.align="center", echo = FALSE}
hfx <- read_rds("../static/selected_weather_day.rds") %>% filter(station_id == 50620, TC_id == "YHZ") %>%
  filter(year == 2019 | year == 2018) %>% filter(!is.na(mean_temp)) %>%
  mutate(year_day = yday(date) + (as.numeric(year)-2018)*365)
p <- hfx %>%
   ggplot(aes(x = year_day, y= mean_temp)) + geom_point() +
   labs(x = "Days since 2018-01-01",
        y = "Mean temperature (°C)",
        title = "Halifax weather from Environment Canada") + my_theme
p
```

---
class: middle
### Fit a cubic spline

```{r, fig.height=5, fig.align="center"}
m1 <- gam( mean_temp ~ s(year_day, bs="cs"), data = hfx)
summary(m1)
```


---
class: middle
### Fit a cubic spline

```{r}
library(broom)
glance(m1)  %>% kable(digits = 2) %>% kable_styling(full_width = FALSE)
tidy(m1) %>% kable(digits = 2) %>% kable_styling(full_width = FALSE)
```

---
class: middle
### Find predicted values and residuals

```{r fig.height=5, fig.align="center"}
library(gratia)
hfx %>% add_fitted(m1) %>% select(year_day, mean_temp, .value) %>%
  add_residuals(m1)  %>% kable()
```

---
class: middle
### Graph "smooth" term

```{r fig.height=5, fig.align="center"}
hfx %>% add_fitted(m1) %>% select(year_day, mean_temp, .value) %>%
  add_residuals(m1) %>%
  ggplot(aes(year_day)) +
  geom_line(aes(y=.value), color="blue", size=2) + 
  geom_point(aes(y = mean_temp)) + my_theme
```


---
class: middle
### Graph "smooth" term

```{r fig.height=5, fig.align="center"}
confint(m1, "s(year_day)", level = 0.95) %>%
  ggplot(aes(year_day)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.25, fill="red") +
  geom_line(aes(y=est), size=2, color="red") + 
  geom_point(aes(y = mean_temp - mean(mean_temp)), data = hfx) + 
  my_theme
```

---
class: middle
### LOESS smooths

```{r, fig.height=5, fig.align="center"}
m2 <- loess( mean_temp ~ year_day, data = hfx)
summary(m2)
```


---
class: middle
###  Predictions

```{r, fig.height=5, fig.align="center"}
hfx %>% select(year_day, mean_temp) %>%
  mutate(prediction = predict(m2),
         residuals = residuals(m2)) %>% kable(digits = 2)
```


---
class: middle
###  Predictions

```{r, fig.height=5, fig.align="center"}
augment(m2, se_fit = TRUE) %>% kable(digits = 2)
```


---
class: middle
###  Residuals

```{r, fig.height=5, fig.align="center"}
augment(m2, se_fit = TRUE) %>% 
  ggplot(aes(x = .resid)) + 
  geom_histogram(binwidth = 1) + my_theme
```

---
class: middle
###  Predictions

```{r fig.height=5, fig.align="center"}
new_data = tibble(year_day = seq(1:730))
augment(m2, newdata = new_data, se_fit = TRUE) %>%
  ggplot(aes(x = year_day, y = .fitted)) + 
  geom_line(size = 2, color="blue") +
  geom_ribbon(aes(ymin = .fitted - 2*.se.fit,
                  ymax = .fitted + 2*.se.fit), alpha = 0.25, fill="blue") +
  geom_point(aes(y = mean_temp), data = hfx) + my_theme
```

---
class: middle
### LOESS with straight lines

```{r fig.height=5, fig.align="center"}
m3 <- loess(mean_temp ~ year_day, data = hfx, 
            degree = 1, span = 0.25)
augment(m3, newdata = new_data, se_fit = TRUE) %>%
  ggplot(aes(x = year_day, y = .fitted)) + 
  geom_line(size = 2, color="blue") +
  geom_ribbon(aes(ymin = .fitted - 2*.se.fit,
                  ymax = .fitted + 2*.se.fit), alpha = 0.25, fill="blue") +
  geom_point(aes(y = mean_temp), data = hfx) + my_theme
```







---
class: middle

# Summary

* Fitting GAM and LOESS smooths

* No model coefficients to extract 

* Computing predicted values with errors and residuals (compared to data)


---
class: middle

# Further reading

* Course notes

* Healy Chapter 6. Work with models. 

---
class: middle, inverse

## Task

Task 10: Fitting and using smoothing models.





