---
title: "Data Visualization"
subtitle: "k-means clustering"
author: "Andrew Irwin, a.irwin@dal.ca"
institute: "Math & Stats, Dalhousie University"
date: "2021-03-05 (updated: `r Sys.Date()`)"
ratio: "16:9"
output:
  xaringan::moon_reader:
    css: [default, hygge, useR-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
library(here)
#library(GGally)
#library(ggfortify)
library(kableExtra)
library(gapminder)
library(broom)
library(plotly)
# library(vegan)
my_theme = theme_classic() + 
  theme(text = element_text(size=18))
```

# Plan

* What is k-means clustering and why do we use it?

* Demonstration 


---
class: middle
### What is k-means?

* A tool to place observations in discrete groups

* Based on a distance calculated on several variables

* You decide how many groups to use (but there is guidance)

* There are variations on the algorithm and other methods (not discussed in this course)

---
## Sample data

```{r}
gapminder_ss <- gapminder %>%
  filter(year == 2007) %>% select(-year)
gapminder_ss %>% select(-country, -continent) %>% arrange(-lifeExp) %>%
  kable() %>% kable_styling(full_width = FALSE)
```


---
class: middle
## Data preprocessing

* Approximately normally distributed, same variance

```{r}
gapminder_scaled <- gapminder_ss %>% 
  mutate(s_logPop = scale(log10(pop)), 
         s_logGDP = scale(log10(gdpPercap)),
         s_lifeExp = scale(lifeExp))
```
```{r echo = FALSE, fig.align='center', out.width="70%", fig.asp = .6}
gapminder_scaled %>% ggplot() +
  geom_density(aes(s_logPop, color = "Population")) + 
  geom_density(aes(s_logGDP, color = "GDP")) + 
  geom_density(aes(s_lifeExp, color = "Life Expectancy")) + theme(aspect.ratio = .6)
```

---
## Perform k-means

```{r}
kclust1 <- kmeans(gapminder_scaled %>%
                    select(s_lifeExp, s_logGDP, s_logPop),
                  centers = 6)
tidy(kclust1) %>% kable()
```


---
## Display clustering

```{r fig.align="center", out.width="60%", fig.asp=0.8}
kclust1 %>% augment(gapminder_scaled) %>%
  ggplot(aes(y = lifeExp, x = log10(pop), color = .cluster, shape = continent)) + 
  geom_point() + theme(aspect.ratio = .8)
```

---
class: middle
## How good is the clustering?

```{r echo = FALSE, out.width="60%", fig.align="center"}
kclusts2 <- 
  tibble(k = 1:9) %>%
  mutate(
    kclust = map(k, ~kmeans(gapminder_scaled %>% select(s_lifeExp, s_logGDP, s_logPop),
                            centers = .x)),
    tidied = map(kclust, tidy),
    glanced = map(kclust, glance),
    augmented = map(kclust, augment, gapminder_scaled)
  )
clusters2    <- kclusts2 %>% unnest(cols = c(tidied))
assignments2 <- kclusts2 %>% unnest(cols = c(augmented))
clusterings2 <- kclusts2 %>% unnest(cols = c(glanced))
clusterings2 %>% mutate(proportion_within = tot.withinss / totss) %>%
  ggplot(aes(x = k, y = proportion_within)) + 
  geom_line() + geom_point() + theme_bw() +
  scale_x_continuous(breaks = 1:9) + ylim(0,1)
```

---
class: middle
# When to use k-means?

* When the goal is to make categories from continuous variables

* Each observation is placed in a cluster

* Based on distance between observations, calculated across all variables

  * Usually best to scale (and maybe transform in other ways) each variable

* Select the number of clusters based on partitioning of "sums of squares"


---
class: middle

# Further reading

* Course notes

---
class: middle, inverse

## Task

* Practice k-means



