---
title: "Data Visualization"
subtitle: "Dynamic graphics"
author: "Andrew Irwin, a.irwin@dal.ca"
institute: "Math & Stats, Dalhousie University"
date: "2021-03-12 (updated: `r Sys.Date()`)"
ratio: "16:9"
output:
  xaringan::moon_reader:
    css: [default, hygge, useR-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
library(here)
library(janitor)
library(plotly)
library(gganimate)
library(palmerpenguins)
my_theme = theme_classic() + 
  theme(text = element_text(size=18))
```

# Plan

* Advantages and disadvantages of dynamic graphics?

* When should you use dynamic graphics?

* Examples

* Application in this course

---
class: middle
### When and Why should you use dynamic graphics?

* Interactivity

* Show changes over time (in the data) over time (as perceived by the viewer)

* Easy to make too complicated

* Requires interaction and may not immediately make the point you want to make

* Distracting

---
class: middle
### Highlights and interaction

[Link to interactive graphic](dynamic-graphics-1.html)

```{r warning=FALSE}
p1 <- penguins %>% plot_ly(x = ~ body_mass_g, y = ~ bill_length_mm) %>%
  add_markers(color = ~ species)
p1
htmlwidgets::saveWidget(as_widget(p1), "dynamic-graphics-1.html")
```

---
class: middle
### Animations

Easy to create with `gganimate`. Make a regular ggplot, then use a variable to show how it changes over time.

```{r warning=FALSE, fig.align="center", out.height="50%", out.width="50%"}
penguins %>% ggplot(aes(bill_length_mm, bill_depth_mm,
                        color = species)) + geom_point() + my_theme
```

---
class: middle
### Animations

Easy to create with `gganimate`. Make a regular ggplot, then use a variable to show how it changes over time.

Shown on next slide.

```{r warning=FALSE}
b = seq(2500, 6500, by = 500)
anim1 <- penguins %>% mutate(size_class = cut(body_mass_g, 
                                              breaks=b, dig.lab=4),
                    group=1:n()) %>%
  ggplot(aes(bill_length_mm, bill_depth_mm,
             color = species, group=group)) + 
  geom_point() + 
  labs(title = "Body mass in the interval {closest_state}") +
  transition_states(size_class) +
  enter_fade() + exit_fade() + my_theme
```

---
class: middle
### Animations

```{r echo=FALSE, warning=FALSE, fig.align="center", fig.asp=1}
animate(anim1)
```

---
class: middle
### Summary

* Dynamic and interactive graphics can be fun to create

* Making good use of these features requires practice

* Use sparingly! Think of your audience and goals

* Best example: animation of gapminder data over years

---
class: middle

# Further reading

* Course notes

---
class: middle, inverse

## Task

* No task for this lesson




