---
title: "Data Visualization"
subtitle: "Making outline maps"
author: "Andrew Irwin, a.irwin@dal.ca"
institute: "Math & Stats, Dalhousie University"
date: "2021-03-15 (updated: `r Sys.Date()`)"
ratio: "16:9"
output:
  xaringan::moon_reader:
    css: [default, hygge, useR-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
library(here)
library(janitor)
library(plotly)
library(gganimate)
library(palmerpenguins)
my_theme = theme_classic() + 
  theme(text = element_text(size=18))
```

# Plan

* Kinds of maps

* Basic maps

* Adding points

* Shading areas

* Projections

---
class: middle
### Map libraries

* `ggmap` (part of tidyverse)

* Many other libraries 
  * `maps`
  * `mapproj` for making projections of the Earth's surface
  * `mapdata` with maps of China and Japan
  * `leaflet` for drawing raster or tiled maps (next lesson)

---
class: middle
### Kinds of maps

* Coastlines and political boundaries

* Natural features (rivers, water bodies)

* Points and filled regions on maps

* How to represent latitude and longitude from a sphere on a screen?

* Tiled maps

---
class: middle
### A basic map

```{r messages=FALSE, warning=FALSE}
library(tidyverse)
WorldData <- map_data('world')  
m1 <- ggplot(WorldData, aes(map_id=region)) + 
        geom_map(map = WorldData,
                 aes(x = long, y = lat),
                 fill = "lightgray", colour = "#7f7f7f", 
                 alpha = 0.5, size=0.5) +
  theme_bw()
```


---
class: middle
### A basic map

```{r echo=FALSE, fig.asp=0.5, out.width="90%", fig.align="center"}
m1
```


---
class: middle
### Colour countries

```{r messages=FALSE, warning=FALSE}
values <- tibble( region = WorldData %>% pull(region) %>% unique())
values <- values %>% mutate(value = runif(nrow(qualities)))
m2 <- ggplot(values, aes(map_id = region)) + 
        geom_map(aes(fill = value),
                 map = WorldData,
                 colour = "lightgray", 
                 alpha = 1, size=0.5) +
   expand_limits(x = c(-180,180), y = c(-80,80))+
  theme_bw()
```

---
class: middle
### Colour countries

```{r out.width="90%", fig.asp = 0.5}
m2
```


---
class: middle
### Show only some countries

```{r messages=FALSE, warning=FALSE}
values <- tibble( region = c("Canada", "China", "Chile"),
                  value  = c(1, 2, 3))
m3 <- ggplot(values, aes(map_id = region)) + 
        geom_map(aes(fill = value),
                 map = WorldData,
                 colour = "lightgray", 
                 alpha = 1, size=0.5) +
   expand_limits(x = c(-180,180), y = c(-80,80))+
  theme_bw()
```

---
class: middle
### Show some countries

```{r out.width="90%", fig.asp = 0.5}
m3
```


---
class: middle
### Colour some countries

```{r messages=FALSE, warning=FALSE}
values <- tibble( region = c("Canada", "China", "Chile"),
                  value  = c(1, 2, 3))
m4 <- ggplot(WorldData, aes(map_id = region)) + 
        geom_map(map = WorldData,
                 colour = "lightgray", fill = "darkgray",
                 alpha = 1, size=0.5) +
        geom_map(aes(fill = value), data = values, map = WorldData) + 
   expand_limits(x = c(-180,180), y = c(-80,80))+
  theme_bw()
```

---
class: middle
### Colour some countries

```{r fig.asp = 0.6, out.width = "90%"}
m4
```


---
class: middle
### Add points

```{r messages=FALSE, warning=FALSE}
library(rnaturalearth)
library(rnaturalearthdata)
library(ggrepel)
canada <- ne_countries(country = "canada", scale = "large", returnclass = "sf")
sites <- tibble(longitude = c(-81, -80), latitude = c(44, 45), label = c("A", "B"))
m5 <- ggplot(data = canada) +
        geom_sf() +
        geom_point(data = sites, 
               aes(x = longitude, y = latitude), 
               size = 4, shape = 21, fill = "darkred") +
  geom_label_repel(data = sites, 
             aes(x = longitude, y = latitude, label = label)) + 
  coord_sf(xlim = c(-140,-50), ylim = c(42,65))
```

---
class: middle
### Add points

```{r fig.asp = 0.6, out.width = "90%"}
m5
```


---
class: middle
### Summary



---
class: middle

# Further reading

* Course notes

---
class: middle, inverse

## Task

* XXXXX




