---
title: "Data Visualization"
subtitle: "Making sliding maps"
author: "Andrew Irwin, a.irwin@dal.ca"
institute: "Math & Stats, Dalhousie University"
date: "2021-03-17 (updated: `r Sys.Date()`)"
ratio: "16:9"
output:
  xaringan::moon_reader:
    css: [default, hygge, useR-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
# html widgets (leaflet) is experimental in xaringan. Try remotes::install_github('yihui/xaringan')
library(tidyverse)
library(leaflet)
library(ggmap)
```

# Plan

* Kinds of maps

* Basic maps

* Adding points

* Shading areas

* Tile service providers

---
class: middle
### Map libraries

* Dynamic: `leaflet` for drawing raster or tiled maps (leafletjs.com)

* Static: `ggmap` (part of tidyverse, has stamen map functions)

---
### A basic map

```{r message=FALSE, warning=FALSE}
m <- leaflet() %>%
  addTiles() %>%  # Add default OpenStreetMap map tiles
  addMarkers(lng=-63.5932, lat=44.63697, popup="Math & Stats, Dalhousie U")
```

---
### A basic map

```{r fig.align="center", out.width="90%", echo=FALSE}
m  # Print the map
```


---
### Adding points

```{r message=FALSE, warning=FALSE}
cities <- read_csv("../static/selected_cities.csv")
m2 <- leaflet(data = cities) %>%
  addTiles() %>%
  addCircleMarkers(~ lng, ~lat, label = ~city)
```

---
### Adding points

```{r fig.align="center", out.width="90%", echo=FALSE}
m2
```

---
### Colour regions

```{r message=FALSE}
library(geojsonio)
pal <- colorNumeric("viridis", NULL)  # make a viridis palette
canada <- geojson_read("https://gist.githubusercontent.com/mikelotis/2156d7c170d10d2c77cb79424fe2137d/raw/7a13748ed7ea5ba64876c77c53b6cb64dd5c3ab0/canada-province.geojson", what="sp")
my_values = runif(13, 0, 10) # random numbers between 0 and 10
m3 <- canada %>% 
  leaflet() %>%
  addTiles() %>%
  addPolygons(fillColor = ~ pal(my_values),
              weight = 1, color = "black") %>%
  addLegend(pal = pal, values = my_values, opacity = 1.0)
```


---
### Colour regions

```{r fig.align="center", out.width="90%", echo=FALSE}
m3
```


---
### Other tiled map services

```{r stamen, cache=TRUE, message=FALSE, fig.align="center", out.width="90%", fig.retina=2}
mymap_terrain <- get_stamenmap(bbox = c(left = -130, bottom = 41, 
                                        right = -50, top = 66), 
                               zoom=5, maptype = "terrain")
m4 <- ggmap(mymap_terrain)  
```
---
### Other tiled map services

```{r message=FALSE, fig.align="center", out.width="90%", fig.retina=2, echo=FALSE}
m4
```

---
### Add text and points

```{r message=FALSE}
my_points <- tibble(lat = c(43+57/60, 49+53/60),
                    lon = c(-59-55/60, -97-9/60),
                    label = c("Sable Is.", "Winnipeg")
)
m5 <- ggmap(mymap_terrain)  +
  geom_point(data = my_points, color="brown") + 
  geom_label(data = my_points, 
            aes(label=label), 
            fill = "#FFFFFF90",
            color = "black", nudge_y = 1.5)
```

---
#### Add text and points

```{r echo=FALSE, fig.align="center", out.width="90%", echo=FALSE, fig.retina=2}
m5
```

---
class: middle
### Summary

* Make a basic map

* Add points and labels

* Fill regions with colour (requires polygons)

* There are several tile services

---
class: middle

# Further reading

* Course notes

---
class: middle, inverse

## Task

* Task 15. Make some maps as described in repository



