---
title: "Data Visualization"
subtitle: "Alternatives to maps"
author: "Andrew Irwin, a.irwin@dal.ca"
institute: "Math & Stats, Dalhousie University"
date: "2021-03-19 (updated: `r Sys.Date()`)"
ratio: "16:9"
output:
  xaringan::moon_reader:
    css: [default, hygge, useR-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
library(ggheatmap)
```

# Plan

* When is a map not appropriate for geospatial data?

* Tile geometry

* Heatmaps with dendrograms

* Specialized heatmap for the USA

---
class: middle
### Why shouldn't I use a map?

* Always think about the purpose

* Is geography the most important feature?

* Are you showing points or areas?

* Will the area of a region complicate interpretation?

* Is the spatial extent one dimensional? What could you do with the other dimension?



---
class: middle
### Summary

* Maps are very effective for showing points where spatial location is the primary information

* To show quantitative data varying along a transcect, a line or dot plot may be better

* Colour shows larger/smaller, positive/negative, but does not show quantitative values well

* Heatmaps (2 dimensional tiles of colours) can be a good alternative


---
### COVID-19 cases & geom_tile

```{r covid-data, message=FALSE, cache=TRUE, echo=FALSE, fig.align="center", out.width="100%", fig.asp = 0.65}
covid <- cansim::get_cansim("13-10-0775-01")
# covid2 <- get_cansim("13-10-0774-01")  # related data with extra information
covid_ss <- covid %>% filter(GEO != "Canada", Gender == "Total, all genders", 
                             `Age group` != "Total, all age groups") %>%
  filter(Statistics == "Community exposures") %>%
  mutate(GEO = str_remove(GEO, " Region"))
covid_ss %>%
  ggplot(aes(y = GEO, x = `Age group`, fill = VALUE)) + 
  geom_tile() +
  theme(axis.text.x = element_text(angle = 90, size=15, hjust = 1, vjust=0.5),
        axis.text.y = element_text(size=14, hjust = 1)) + 
  labs(x = "", y="")
```


---
### Same data with ggheatmap

```{r message=FALSE, echo=FALSE, fig.align="center", out.width="90%"}
covid_ss4 <- covid_ss %>% select(GEO, `Age group`, VALUE) %>%
  pivot_wider(names_from = GEO, values_from = VALUE) 
data.frame(covid_ss4 %>% select(-`Age group`), 
              row.names = covid_ss4 %>% pull(`Age group`)) %>%
  ggheatmap()
```


---
### Scale columns to have sum 1

```{r message=FALSE, echo=FALSE, fig.align="center", out.width="90%"}
data.frame(covid_ss4 %>% select(-`Age group`), 
              row.names = covid_ss4 %>% pull(`Age group`)) %>%
  scale(center = FALSE, scale = colSums(.)) %>%
  as.data.frame() %>%
  ggheatmap()
```



---
### Specialized maps of USA

```{r}
library(statebins)
load("../static/election.rda")
m4 <- election %>% ggplot(aes(state = state, fill = pct_trump)) +
  geom_statebins() +
  theme_statebins() +
  labs(fill="Percent Trump")
```

---
### Specialized maps of USA

```{r fig.align=TRUE, out.width="90%", echo=FALSE, fig.asp=0.75}
m4
```


---
class: middle

# Further reading

* Course notes

* Healy Chapter 7

---
class: middle, inverse

## Task

* Task 16 as described in repository



