---
title: "Data Visualization"
subtitle: "Text, Factors, Dates, and Times"
author: "Andrew Irwin, a.irwin@dal.ca"
institute: "Math & Stats, Dalhousie University"
date: "2021-03-22 (updated: `r Sys.Date()`)"
ratio: "16:9"
output:
  xaringan::moon_reader:
    css: [default, hygge, useR-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
library(forcats)
library(lubridate)
library(palmerpenguins)
library(kableExtra)
library(glue)
library(unglue)
my_theme = theme_bw() + theme(text = element_text(size=25))
```

# Plan

* Tools to help you work with data that are not numbers

* Specialized functions for working with strings, factors and dates

* Strings package `stringr`, `glue`, `unglue` and functions `str_squish`, `glue` and more

* Factors package `forcats` and functions `fct_*`

* Date and time package: `lubridate` and functions `ymd`, `ymdhms`, `yday`, `decimal_date` and more


---
class: middle
### Challenges that arise with text

* Extra spaces

* Upper and lower case differences

* Locales (non-English text); see [ragg](https://www.tidyverse.org/blog/2021/02/modern-text-features/) for plotting with non-Latin text

* Difference between factors and strings

---
class: middle
### Extra spaces

Data entered by a human in a web form or spreadsheet often has inconspicuous spaces, for example after the last letter, or two spaces between a word. This is easily ignored when read by humans, but creates havoc for computers.  `str_squish` gets rid of these troublesome spaces.

```{r}
str_squish(" a crazy sentence   with 
           too many spaces in    strange places.     ")
```

---
class: middle
### Letter case

Upper and lower case differences are often not noticed by humans, but matter to a computer. One solution is to convert all text to upper, lower, or title case.

```{r}
tibble(original = c("Apple", "apple", "APPLE", "aPpLe"),
       lower = str_to_lower(original),
       upper = str_to_upper(original),
       title = str_to_title(original)) %>% kable()
```

---
### Getting data in and out of strings

```{r}
tibble(name = c("Andrew", "Susan", "Yong"),
       age = c(12, 21, 35)) %>%
  mutate(sentence = glue("{name} is {age} years old.")) %>% kable()
```

---
### Getting data in and out of strings

```{r include = FALSE}
t0 <- tibble(name = c("Andrew", "Susan", "Yong"),
       age = c(12, 21, 35)) %>%
  mutate(sentence = glue("{name} is {age} years old."))
```

```{r}
t0 %>% select(sentence) %>%
  mutate( unglue_data(sentence, "{name} is {age} years old.")) %>%
  mutate( age = as.numeric(age),
          next_year = age + 1) %>% kable()
```

---
class: middle
### Challenges that arise with factors

* Plot order on scales (axes, color scale)

* Too many factors

* Mapping colours to specific levels (later lesson on colour)


---
class: middle
### Plot order

Here's a plot of mean penguin body mass by species and sex. What's the order?

```{r warning=FALSE, message=FALSE, fig.asp = 0.5, fig.align="center", out.width="90%"}
penguins %>% ggplot(aes(x = body_mass_g, y = island)) + 
  stat_summary() +
  facet_wrap(~ sex) + theme_bw() 
```


---
class: middle
### Plot order

Order from smallest to largest, top to bottom. Watch out for NAs.

```{r warning=FALSE, message=FALSE, out.width = "50%", fig.align="center", fig.asp =0.75}
penguins %>% na.omit() %>%
  ggplot(aes(x = bill_length_mm, 
                   y = fct_reorder(species, bill_length_mm,
                                   .desc=TRUE))) + 
  stat_summary() + my_theme
```


---
class: middle
### Custom order

Order from smallest to largest, top to bottom. Watch out for NAs.

```{r warning=FALSE, message=FALSE, out.width = "50%", fig.align="center", fig.asp =0.75}
penguins %>%
  ggplot(aes(x = bill_length_mm, 
             y = fct_relevel(species, "Gentoo", "Adelie"))) + 
  stat_summary() + my_theme
```

---
class: middle
### Challenges that arise with dates and times

* Date format

* Extracting components of date or time

* Formatting axes on plots

* Arithmetic with dates and times

* Time zones

---
### Converting text to dates

```{r}
tibble(date = c("01/02/03", "121006", "05/12/08", "11-03-21"),
       ymd = ymd(date),
       dmy = dmy(date),
       mdy = mdy(date)) %>% kable()
```


---
### Dates and times

```{r warning=FALSE, message=FALSE, out.width = "50%", fig.align="center", fig.asp =0.75}
tibble(date = c("2021/03/11 10:05", "2021/03/12 15:12", 
                "2021/03/11 15:14", "2021/03/11 11:50 PM"),
       dt = ymd_hm(date),
       value = 1:4) %>%
  ggplot(aes(y = dt, x = value)) + geom_point(size=4) + my_theme
```

---
### Dates and times

```{r warning=FALSE, message=FALSE, out.width = "50%", fig.align="center", fig.asp =0.75}
tibble(date = c("2021/03/11 10:05", "2021/03/12 15:12", 
                "2021/03/11 15:14", "2021/03/11 11:50 PM"),
       dt = ymd_hm(date),
       value = 1:4) %>%
  ggplot(aes(y = dt, x = value)) + geom_point(size=4) + 
  scale_y_datetime(date_labels = "%H:%M") + my_theme
```

---
### Time arithmetic

```{r}
tibble(date = c("2021/03/11 10:05", "2021/03/12 15:12", 
                "2021/03/11 15:14", "2021/03/11 11:50 PM"),
       dt = ymd_hm(date),
       value = 1:4) %>%
  mutate(elapsed = dt - min(dt)) %>% kable()
```

---
### Time arithmetic

```{r include=FALSE}
t1 <- tibble(date = c("2021/03/11 10:05", "2021/03/12 15:12", 
                "2021/03/11 15:14", "2021/03/11 11:50 PM"),
       dt = ymd_hm(date),
       value = 1:4) 
```

```{r warning=FALSE, message=FALSE, out.width = "50%", fig.align="center"}
t1 %>% mutate(elapsed = dt - min(dt)) %>%
  ggplot(aes(x = as.numeric(elapsed)/3600, 
             y = value)) + geom_point(size=4) +
  labs(x = "Time in hours since start") + my_theme
```

---
class: middle

# Further reading

* Course notes

* Healy appendix

* R4DS Chapter 14 [Strings](https://r4ds.had.co.nz/strings.html), Chapter 15 [Factors](https://r4ds.had.co.nz/factors.html), and Chapter 16 [Dates and times](https://r4ds.had.co.nz/dates-and-times.html)

---
class: middle, inverse

## Task

* Task 17 as described in repository



