---
title: "Data Visualization"
subtitle: "Colour"
author: "Andrew Irwin, a.irwin@dal.ca"
institute: "Math & Stats, Dalhousie University"
date: "2021-03-24 (updated: `r Sys.Date()`)"
ratio: "16:9"
output:
  xaringan::moon_reader:
    css: [default, hygge, useR-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
library(forcats)
library(lubridate)
library(palmerpenguins)
library(kableExtra)
library(paletteer)
library(gapminder)
my_theme = theme_bw() + theme(text = element_text(size=25))
knitr::opts_chunk$set(
  fig.width = 6, fig.asp = 0.65, fig.align="center", out.width = '70%', dev='svglite'
)
```

# Plan

* Discrete colour scales

* Continuous colour scales

* viridis colour

* other palettes

* custom colours

---
class: middle
### Colour guidelines - Discrete

* Not too many

* Keep brightness constant

---
class: middle
### Colour guidelines - Continuous

* One shade, vary brightness

* Two shades if there is a natural middle or 0, white in middle

* Avoid red-green diverging scales


---
class: middle
### Discrete colour scales

```{r warning=FALSE, fig.asp = 0.7, out.width="60%", fig.align="center"}
penguins %>% ggplot(aes(body_mass_g, flipper_length_mm,
                        color = species)) + geom_point() + my_theme
```

---
class: middle
### Discrete colour scales

```{r warning=FALSE, fig.asp = 0.7, out.width="60%", fig.align="center"}
penguins %>% ggplot(aes(body_mass_g, flipper_length_mm,
                        color = species)) + geom_point() + my_theme +
  scale_color_viridis_d(end = 0.7) 
```


---
class: middle
### Discrete colour scales

```{r warning=FALSE, fig.asp = 0.7, out.width="60%", fig.align="center"}
penguins %>% ggplot(aes(body_mass_g, flipper_length_mm,
                        color = species)) + geom_point() + my_theme +
  scale_color_paletteer_d("ggthemes::hc_default")
```


---
class: middle
### Discrete colour scales

```{r warning=FALSE, fig.asp = 0.7, out.width="60%", fig.align="center"}
penguins %>% ggplot(aes(body_mass_g, flipper_length_mm,
                        color = species)) + geom_point() + my_theme +
  scale_color_manual(values = c("red", "blue", "orange"))
```

---
### Continuous colour shades

```{r warning=FALSE, fig.asp = 0.7, out.width="60%", fig.align="center"}
gapminder %>% filter(country %in% c("Canada", "China", "Chile", "Iceland")) %>%
  ggplot(aes(year, lifeExp, color = log10(pop))) +
  geom_point(size=3) + my_theme
```


---
### Continuous colour shades - diverging

```{r warning=FALSE}
p1 <- penguins %>% mutate(body_mass_centered = 
                      body_mass_g - mean(body_mass_g, na.rm=TRUE)) %>%
  ggplot(aes(flipper_length_mm, bill_length_mm, 
             color = body_mass_centered)) +
  geom_point(size=3) + theme_bw() +
  # scale_color_distiller(type="div", palette="RdBu", limits = c(-2000,2000)) 
  scale_color_fermenter(type="div", 
                        palette="RdBu", 
                        limits = c(-2000,2000)) 
```


---
### Continuous colour shades - diverging

```{r echo=FALSE, warning=FALSE, fig.asp = 0.7, out.width="80%", fig.align="center"}
p1 + my_theme +
  labs(color = "Mass relative\nto mean (g)")
```

---
class: middle

# Further reading

* Course notes

* Healy and Wilke sections on colour


---
class: middle, inverse

## Task

* Practice using colour

* No assigned task



