---
title: "Data Visualization"
subtitle: "Theme"
author: "Andrew Irwin, a.irwin@dal.ca"
institute: "Math & Stats, Dalhousie University"
date: "2021-03-26 (updated: `r Sys.Date()`)"
ratio: "16:9"
output:
  xaringan::moon_reader:
    css: [default, hygge, useR-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
library(forcats)
library(lubridate)
library(palmerpenguins)
library(kableExtra)
library(paletteer)
library(gapminder)
my_theme = theme_bw() + theme(text = element_text(size=25))
dpi = 72
knitr::opts_chunk$set(
  fig.width = 6, fig.asp = 0.65, fig.align="center", out.width = '70%', dev='svglite'
)
```

# Plan

* What is the `theme` of a visualization?

* Ready-to-use themes

* Titles

* Text size, colour, bold, italics, and more

* Spacing and positioning


---
### Starting figure

```{r echo=FALSE}
p1 <- gapminder %>% # filter(continent == "Europe") %>%
  ggplot(aes(x = year, y = lifeExp)) + 
  # geom_line(aes(group=country)) + 
  geom_smooth(method = "loess", formula = y ~ x, aes(color = continent))
p1
```

---
### Ready to use themes

```{r}
p1 + theme_bw()
```

---
### Ready to use themes

```{r}
p1 + theme_classic()
```
---
### More themes

* `theme_dark`, `theme_gray`, `theme_light`

* From `ggthemes`: `theme_wsj`, `theme_tufte`, `theme_fivethirtyeight`, `theme_economist`, `theme_clean`


---
### All text larger

```{r}
p1 + theme(text = element_text(size=22))
```

---
### Sizes for different elements

```{r}
p1 + theme(axis.text.x = element_text(size=16),
           axis.title.y = element_text(size=18),
           legend.title = element_text(size=16),
           legend.text = element_text(size=18),
           plot.title = element_text(size=21),
           plot.subtitle = element_text(size=20)) +
  labs(x = "Year", 
       y = "Life Expectancy at birth",
       title = "Life expectancy averaged by continent",
       subtitle = "From Gapminder.org")
```

---
### Other features of text

```{r}
p1 + labs(title = "Average life expectancy") + 
  theme(plot.title = element_text(family = "serif",
                                  face = "italic",
                                  color = "blue",
                                  size = 30,
                                  hjust = 1, 
                                  lineheight = 2))
```

---
### Fancy text: ggtext

```{r}
library(ggtext)
p1 + labs(title = "**Gapminder** data for _life expectancy_<sup>1</sup>",
          caption = "<sup>1</sup>from gapminder.org") +
  theme(plot.title = element_markdown(size = 24),
        plot.caption = element_markdown(size = 13))
```

---
### Move and redesign the legend

```{r}
p1 + theme(legend.position = "top",
           legend.direction = "horizontal",
           legend.text = element_text(size = 11, vjust = 0.5),
           legend.title = element_text(size = 12, vjust = 0.5),
           legend.key.size = unit(0.9, "cm"))
```

---
### Move and redesign the legend

```{r}
p1 + theme_bw() + theme(legend.position = c(0.03, 0.98),
                        legend.justification = c(0,1),
           legend.direction = "vertical",
           legend.background = element_rect(fill = "transparent"),
           legend.text = element_text(size = 10, vjust = 0.5),
           legend.title = element_blank(),
           legend.key.size = unit(0.8, "cm")) + 
  guides(color = guide_legend(ncol = 3))

```

---
### Formatting numbers on scales

```{r}
p1 + scale_x_continuous(labels = scales::dollar) +
  scale_y_continuous(labels = scales::percent) + # also comma
  theme(axis.text = element_text(size=14))
```

---
class: middle

# Further reading

* Course notes

* Healy and Wilke sections on colour


---
class: middle, inverse

## Task

* Practice using colour

* No assigned task



