---
title: "Data Visualization"
subtitle: "Graphics output"
author: "Andrew Irwin, a.irwin@dal.ca"
institute: "Math & Stats, Dalhousie University"
date: "2021-03-29 (updated: `r Sys.Date()`)"
ratio: "16:9"
output:
  xaringan::moon_reader:
    css: [default, hygge, useR-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
library(forcats)
library(lubridate)
library(palmerpenguins)
library(kableExtra)
library(paletteer)
library(gapminder)
my_theme = theme_bw() + theme(text = element_text(size=25))
knitr::opts_chunk$set(
  echo = FALSE,
  fig.width = 6, fig.asp = 0.65, fig.align="center", out.width = '70%', dev='svglite'
)
```

# Plan

* Kinds of graphics 

* How to describe the size of a graphic

* Knitted documents and separate files


---
### Starting figure

```{r}
p1 <- gapminder %>% # filter(continent == "Europe") %>%
  ggplot(aes(x = year, y = lifeExp)) + 
  # geom_line(aes(group=country)) + 
  geom_smooth(method = "loess", formula = y ~ x, aes(color = continent)) +
  theme_bw() + 
  theme(text = element_text(size=20))
p1
```

---
### Kinds of graphics

* Vector graphics (pdf, svg)

* Bitmapped or raster graphics (png, jpeg)

* Why does it matter?
  
  * File sizes
  
  * Quality when printed
  
  * Appearance on screen
  
---
### Overview

* Device types: png, ragg, jpeg, svg, pdf

* Size: fig.height, fig.width (in inches, default 7), and fig.asp (aspect ratio: height / width)

* Display size: out.width, out.height (in percentage of display area, e.g., '70%')

* Resolution:  dpi (dots per inch; 96-250 on screen, 300-600 when printed)

* Alignment: fig.align = left, right, center

---
### PNG example

Chunk options: `fig.width = 6, fig.asp = 0.65, fig.align="center", out.width = '70%', dev = 'png'`

```{r dev='png'}
p1 
```

---
### Higher resolution PNG example

Chunk options: `dev='png', dpi=150`

```{r dev='png', dpi=150}
p1 
```

---
### RAGG example

Chunk options: `dev='ragg_png', dpi = 150`

```{r dev='ragg_png', dpi = 150}
p1 
```


---
### JPEG example

Chunk options: `dev='jpeg', dpi = 150`

```{r dev='jpeg', dpi = 150}
p1 
```

---
### SVG example

Chunk options: `dev='svglite'`

```{r dev='svglite'}
p1 
```

---
### PDF example

The file looks great on [my computer](33-graphics-output_files/figure-html/pdfoutput-1.pdf) (look in `XXXX_files` folder), but is not shown properly in a web page.

Chunk options: `dev='pdf'`

```{r pdfoutput, dev='pdf'}
p1 
```


---
class: middle

# Further reading

* Course notes


---
class: middle, inverse

## Task

* No assigned task



