---
title: "Live coding 11"
author: "Andrew Irwin"
date: "2021-03-25 8:35"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

```

# Agenda

* Questions arising from lessons, tasks
* Working with strings, factors, dates and times

The plan for the live coding is for us to work together to solve problems using tools from this week's lessons. 


## Strings

* String clean up with str_squish, str_lo_lower, str_remove, str_remove_all

```{r}

```

* Building strings from data with glue

```{r}

```

* Extracting data from text using unglue

```{r}

```


## Factors

* Reordering factors using a quantitative variable

```{r}

```

* Reordering a factor using an arbitrary order

```{r}


```

* Grouping several factors together

```{r}


```

## Dates and times

* Converting text to dates and dates/times


```{r}

```

* Arithmetic with date/times (subtracting, etc)

```{r}

```

* Formatting axis labels on plots

```{r}

```
