---
title: "Live coding 10"
author: "Andrew Irwin"
date: "2021-03-18 8:35"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggmap)
library(mapproj)
library(leaflet)
library(statebins)

```

# Agenda

* Questions arising from lessons, tasks
* Maps and alternatives

The plan for the live coding is for us to work together to solve problems using tools from this week's lessons. 

## Outline maps

* Picking a country or several from the world map

```{r}

```

* Drawing the map

```{r}

```

* Filling countries with colours

```{r}

```

* Adding points to a map

```{r}

```

* Making a projection

```{r}

```

## Sliding maps

* Simple sliding map with leaflet

```{r}

```

* Setting latitude, longitude, zoom level with `setView`

```{r}


```

* Adding points to the map

```{r}


```

## Alternatives to maps

* geom_tile for a "heatmap"


```{r}

```

* statebins (see homework for a global version)

```{r}

```

