---
title: "Live coding 10"
author: "Andrew Irwin"
date: "2021-03-18 8:35"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggmap)
library(mapproj)
library(leaflet)
library(statebins)
```

# Agenda

* Questions arising from lessons, tasks
* Maps and alternatives

The plan for the live coding is for us to work together to solve problems using tools from this week's lessons. 

## Outline maps

* Picking a country or several from the world map

```{r}
m_data <- map_data("world")
m_data
```

* Drawing the map

```{r}
m1 <- ggplot() +
  geom_map(aes(map_id = region), map = m_data, data = m_data) +
  expand_limits(x = c(-180, 180), y = c(-90,90))
m1
```

* Filling countries with colours

```{r}
m2 <- ggplot() + 
  geom_map(aes(map_id = region, fill=region), map = m_data, data = m_data,
           show.legend = FALSE) + 
  expand_limits(x = c(-180, 180), y = c(-90,90))
m2
```

* Colour only some countries

```{r}
selected_countries <- c("Canada", "France", "USA")
m3 <- ggplot() + 
  geom_map(aes(map_id = region), map = m_data, data = m_data) + 
  geom_map(aes(fill = region, map_id = region), map = m_data,
           data = m_data %>% filter(region %in% selected_countries)) +
  expand_limits(x = c(-180, 180), y = c(-90,90))
m3
```
)
* Adding points to a map

Here are some data I wanted to add to a map yesterday.

```{r}
ts <- tribble(
  ~ name, ~ lat, ~ long,
  "Cariaco", 10.50, -64.66,
  "L4", 50+15/60, -4-13/60,
  "Port Hacking", -34-4/60, 151+8/60,
  "Gullmar", 58+15/60, 11+27/60,
)
library(ggrepel)
m1 + 
  geom_point(aes(long, lat), color="green", size=3, data = ts) +
  geom_label_repel(aes(long, lat, label = name), data = ts)
```

* Making a projection

Show greenland on a polar projection.

```{r}
greenland <- map_data("world", c("Greenland", "Iceland", "Canada"))
ggplot() + 
  geom_map(aes(map_id = region), map = greenland, data = greenland) + 
  expand_limits(x = c(-80, -20), y = c(60, 85)) + 
  coord_map("azequalarea")   # ?mapproject
```

## Sliding maps

* Simple sliding map with leaflet

```{r}
leaflet() %>% addTiles()
```

* Setting latitude, longitude, zoom level with `setView`

```{r}
leaflet() %>% addTiles() %>% setView(0,0,5)
```

* Adding points to the map

```{r}
leaflet() %>% addTiles() %>% setView(0,0,2) %>%
  addMarkers(ts$long, ts$lat, popup = ts$name)
leaflet() %>% addTiles() %>% setView(0,0,2) %>%
  addCircleMarkers(ts$long, ts$lat, popup = ts$name)

```

## Alternatives to maps

* geom_tile for a "heatmap"

```{r}
t1 <- expand.grid(region = c("A", "B", "C", "D", "Z"),
                  month = c("Jan", "Feb", "Mar"))
t1 %>% mutate(value = 1:15) %>%
  ggplot() +
  geom_tile(aes(x = region, y = month, fill = value))
```

* statebins (see homework for a global version)

```{r}
t2 <- tibble(
  state = c("AZ", "CA", "NM", "RI", "NJ", "FL", "NJ"),
  value = 1:7
)
t2 %>% 
  ggplot() +
  geom_statebins(aes(state = state, fill = value))
```

State area

```{r}
t3<- tibble(state = state.abb,
            area = state.area)
ggplot() + 
  geom_statebins(aes(state = state, fill = log10(area)), data = t3)
```

