---
title: "Weather calculations"
author: "Andrew Irwin"
date: "11/01/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
t_day <- readRDS("../static/selected_weather_day.rds")
```


## Today in history

* compute quantiles of minimum and maximum temperature
* compute fraction of days with precipitation, and amounts of precipitation.
* compute proportion of runs of 3 days with 0, 1, 2, 3 days of preciptitation. are they independent?
* compute distribution of changes between today and tomorrow in min, max temperature; presumably symmetric or nearly so, so also mean absolute difference
* amount of snow on ground
* sunrise, sunset, max sun angle, moon phase

* plot Atlantic ocean temperature, heat delivered, air temperature

## today and this week

probability of rain, snow, amount of snow on ground.

```{r}
todayFilter <- function(date = now(), df = t_day, week=FALSE) {
   m <- month(date)
   d <- day(date)
   w <- week(date)
   doy <- yday(date)
   if (week) {
      df %>% filter(week(date) == w)
   } else {
     df %>% filter(yday(date) == doy)
   }
}
todayFilter()
median_gt_0 <- function(x) median(x[x>0], na.rm=TRUE)
snow <- function(df = t_day) {
  df %>% summarize(n = n(),
                   snow_ground = median_gt_0(snow_grnd),
                   snow_cm = median_gt_0(total_snow),
                   snow_y = mean(total_snow > 0, na.rm=TRUE),
                   rain_y = mean(total_rain > 0, na.rm=TRUE)
  )
}
snow(todayFilter(week=TRUE) %>% group_by(TC_id))
```

## Change over time (same day, differnet years)

## climatologies
